% !TEX root = thesis.tex

\chapter{A case competition typology}\label{ch:case-competition-typology}

In Part \ref{part:case-facts} of this dissertation, I discussed a first aspect of case competition in headless relatives. There is a fixed scale that determines which case wins the case competition. This is the same case scale crosslinguistically. I repeat the case scale from Chapter \ref{ch:recurring} in \ref{ex:case-scale-two-patterns}.

\ex. \ac{nom} < \ac{acc} < \ac{dat}\label{ex:case-scale-two-patterns}

In Chapter \ref{ch:decomposition} within Part \ref{part:case-facts}, I argued that a cumulative case decomposition can derive the case scale. This does not only hold for case competition in headless relatives, but also for syncretism patterns and morphological case containment patterns. In a cumulative case composition, the scale in \ref{ex:case-scale-two-patterns} can be interpreted as follows: the accusative contains all features the nominative contains plus one more. Similarly, the dative contains all features the accusative contains plus one. Therefore, the dative can be considered more complex than the accusative, and the accusative more complex than the nominative. In line with that, I refer to cases more to the right on the case scale as being more complex cases than cases more to the left on the scale.

This part of the dissertation, Part \ref{part:variation}, focuses on a second aspect to headless relatives. This aspect is not stable crosslinguistically, but it differs across languages. Languages differ in whether they allow the internal case (the case from the relative clause) and the external case (the case from the main clause) to surface when either of them wins the case competition. Metaphorically speaking, even though a case wins the case competition, it is a second matter whether it is allowed to come forward as a winner. Four patterns are logically possible for languages: (1) the internal case and the external case are allowed to surface, (2) only the internal case is allowed to surface, and the external case is not, (3) only the external case is allowed to surface, and the internal case is not, (4) neither the internal nor the external case is allowed to surface.\footnote{
On the surface, the last pattern cannot be distinguished from a language that does not have case competition and does not allow for any case mismatches. I come back to this matter in \ref{sec:possible-patterns}, where I argue that there actually is case competition in play.
}
I show in this chapter that one of these logically possible patterns is not attested in any natural language.

In this dissertation I discuss languages of which headless relatives have been described in the literature. As I write about case competition, I only focus on languages that morphologically distinguish between case, specifically the nominative, the accusative and the dative. By no means do I claim that my language sample is representative for the languages of the world. However, they build on independently and strongly established facts, which are the case scale from Chapter \ref{ch:recurring} and the subset requirement of the external head, to be discussed in Chapter \ref{ch:relativization}. Therefore, I predict that my generalizations hold for all natural languages.

The next section introduces the patterns that are logically possible with case competition. In Section \ref{sec:pattern-i} to Section \ref{sec:pattern-iv}, I discuss the patterns one by one, and I give examples when the pattern is attested. In Section \ref{sec:without-case-competition}, I make a sidestep to languages that do not show any case competition, and I give a typology of headless relatives.


\section{Four possible patterns}\label{sec:possible-patterns}

Case competition has two aspects. The first aspect is the topic of Part \ref{part:case-facts} of the dissertation. It concerns which case wins the case competition. This is decided by the same case scale for all languages. The second aspects is the topic of Part \ref{part:variation} of the dissertation. This one concerns whether the case that wins the case competition is actually allowed to surface. It namely differs per language whether it allows the internal or the external case to do so.

Metaphorically, the second aspect can be described as a language-specific approval committee. The committee learns (from the first aspect) which case wins the case competition. Then it can either approve this case or not approve it. This approval happens based on where the winning case comes from: from inside to the relative clause (internal) or from outside to the relative clause (external). It is determined per language whether it approves the internal case, the external case, both of them or none of them. The approval committee can only approve the winner of the competition or deny it, it cannot propose an alternative winner. In this metaphor, the approval of the committee means that a particular case is allowed to surface. When the case is not allowed to surface, the headless relative as a whole is ungrammatical.

Taking this all together, there are four patterns possible in languages. First, the internal case and the external case are allowed to surface. Second, only the internal case is allowed to surface, and the external case is not. Third, only the external case is allowed to surface, and the internal case is not. Fourth, neither the internal nor the external case is allowed to surface. In what follows, I introduce these four possible patterns.

The first possible pattern is that of a language that allows the internal case and the external case to surface when either of them wins the case competition. This pattern might look familiar, because it is the one that Gothic has, which I discussed in Chapter \ref{ch:recurring}. Table \ref{tbl:case-competition-int-ext} (repeated from Table \ref{tbl:summary-gothic-simple}) illustrates what the pattern for such a language looks like.

The left column shows the internal case between square brackets. The upper row shows the external case between square brackets. The other cells indicate the case of the relative pronoun.
The top-left to bottom-right diagonal corresponds to the examples in which the internal and external case match. The three cells in the bottom-left corner, marked in light gray, are the situations in which the internal case surfaces when it wins the competition. The three cells in the top-right corner, marked in dark gray, are the situations in which the external case surfaces when it wins the competition.
All these instances are grammatical.

\begin{table}[H]
  \center
  \caption{Internal and external case allowed}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \cellcolor{DG}\ac{acc}
           & \cellcolor{DG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \cellcolor{LG}\ac{acc}
           & \ac{acc}
           & \cellcolor{DG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \cellcolor{LG}\ac{dat}
           & \cellcolor{LG}\ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-int-ext}
\end{table}

The second possible pattern is that of a language that allows the internal case to surface when it wins the case competition, but it does not allow the external case to do so. In this type of language, the internal case gets to surface when it is more complex than the external one. When the external case is more complex, it is not allowed to surface, and the headless relative construction is ungrammatical.

Table \ref{tbl:case-competition-only-int} illustrates what the pattern for such a language looks like. Compared to the first pattern, it has three cells in which there is no grammatical relative pronoun.
The top-left to bottom-right diagonal corresponds to the examples in which the internal and external case match.
The three cells in the bottom-left corner, marked in light gray, are the situations in which the internal case surfaces when it wins the competition.
Just as in the first pattern, these six instances are grammatical.
The three cells in the top-right corner, marked in dark gray, are the situations in which the external case surfaces when it wins the competition. These instances are not grammatical for this type of language. The reasoning behind that is that the language does not allow the external case to surface when it wins the case competition.

\begin{table}[H]
  \center
  \caption{Only internal case allowed}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \cellcolor{DG}*
           & \cellcolor{DG}*
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \cellcolor{LG}\ac{acc}
           & \ac{acc}
           & \cellcolor{DG}*
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \cellcolor{LG}\ac{dat}
           & \cellcolor{LG}\ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-only-int}
\end{table}

The third possible pattern is that of a language that allows the external case to surface when it wins the case competition, but it does not allow the internal case to do so. In this type of language, only the external case gets to surface when it is more complex. When the internal case is more complex, it is not allowed to surface, and the headless relative construction is ungrammatical.

Table \ref{tbl:case-competition-only-ext} illustrates what the pattern for such a language looks like. Comparing this pattern to the second one, the ungrammatical cells are here the three on the other side of the diagonal.
The top-left to bottom-right diagonal corresponds to the examples in which the internal and external case match. Just as in the first and second pattern, these instances are grammatical.
The three cells in the bottom-left corner, marked in light gray, are the situations in which the internal case surfaces when it wins the competition. Unlike in the first and the second pattern, these instances are not grammatical for this type of language. The reasoning behind that is that the language does not allow the internal case to surface when it wins the case competition.
The three cells in the top-right corner, marked in dark gray, are the situations in which the external case surfaces when it wins the competition. Just as in the first pattern but unlike in the second pattern, these instances are grammatical.

\begin{table}[H]
  \center
  \caption{Only external case allowed}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \cellcolor{DG}\ac{acc}
           & \cellcolor{DG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \cellcolor{LG}*
           & \ac{acc}
           & \cellcolor{DG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \cellcolor{LG}*
           & \cellcolor{LG}*
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-only-ext}
\end{table}

The fourth possible pattern is that of a language that allows neither the internal nor the external case to surface when either of them wins the case competition. When the internal case or the external is more complex, it is not allowed to surface, and the headless relative construction is ungrammatical. In other words, when the internal and the external case differ, there is no grammatical headless relative construction possible.

Table \ref{tbl:case-competition-none} illustrates what the pattern for such a language looks like.
The top-left to bottom-right diagonal corresponds to the examples in which the internal and external case match. Just as in the first, second and third pattern, these instances are grammatical.
The three cells in the bottom-left corner, marked in light gray, are the situations in which the internal case surfaces when it wins the competition. Just as the third pattern, but unlike the first and the second pattern, these instances are not grammatical for this type of language.
The three cells in the top-right corner, marked in dark gray, are the situations in which the external case surfaces when it wins the competition. Just as the second pattern, but unlike the first and the third pattern, these instances are not grammatical for this type of language. The reasoning behind the ungrammaticality of these six cells is that the language allow neither the internal case nor the external case to surface when either of them wins the case competition.

On the surface, this pattern cannot be distinguished from a pattern that does not have case competition and does not allow for any case mismatches. I understand `a language with case competition' as a language that compares the internal and external case in its headless relatives. If the internal and external case are not compared in this type of language, it would be unclear why the diagonal is different from all the other cells. The source of ungrammaticality for the cells in Table \ref{tbl:case-competition-none} can only come from the comparing the internal and external case and concluding that the internal case and the external case differ. The grammaticality of the diagonal follows from the conclusion that the internal and the external case match. In Section \ref{sec:without-case-competition} I discuss languages in which the internal and external case are not compared to each other.

\begin{table}[H]
  \center
  \caption{Neither internal nor external allowed}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \cellcolor{DG}*
           & \cellcolor{DG}*
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \cellcolor{LG}*
           & \ac{acc}
           & \cellcolor{DG}*
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \cellcolor{LG}*
           & \cellcolor{LG}*
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-none}
\end{table}

In this chapter I show that three of these four patterns I introduced are attested crosslinguistically. Section \ref{sec:pattern-i} shows that the first pattern, in which either the internal case or the external case can surface, is exemplified by Gothic (repeated from Chapter \ref{ch:recurring}) and by Old High German. The second pattern, in which only the internal case can surface, is illustrated by Modern German in Section \ref{sec:pattern-ii}. To my knowledge, there is no language in which only the external case can surface when it wins the case competition. This is discussed in \ref{sec:pattern-iii}. Section \ref{sec:pattern-iv} shows a language that allows neither the internal nor the external case to surface when either of them wins the competition, namely Polish.


\section{Internal and external case allowed}\label{sec:pattern-i}

This section discusses the situation in which the internal case and the external case are allowed to surface when either of them wins the case competition. I repeat the pattern from Section \ref{sec:possible-patterns} in Table \ref{tbl:case-competition-int-ext-repeated}.

\begin{table}[H]
  \center
  \caption{Internal and external case allowed (repeated)}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{acc}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \ac{dat}
           & \ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-int-ext-repeated}
\end{table}

Two examples of languages that show this pattern are Gothic and Old High German. In this section, I repeat the summary of the findings from Gothic (from Chapter \ref{ch:recurring}), and I present the data for Old High German, which is the result of my own research.

In Chapter \ref{ch:recurring}, I discussed case competition in Gothic headless relatives, based on the work of \citet{harbert1978}. I repeat the results from Section \ref{sec:pattern-rels} in Table \ref{tbl:summary-gothic-repeated}.
In Gothic, the relative pronoun is allowed to surface in the internal case and the external case. The top-left to bottom-right diagonal corresponds to the examples in which the internal and external case match. The three cells in the bottom-left corner, marked in light gray, are the situations in which the internal case surfaces when it wins the competition. The three cells in the top-right corner, marked in dark gray, are the situations in which the external case surfaces when it wins the competition.
 All these instances are grammatical. The examples corresponding to the cells in Table \ref{tbl:summary-gothic-repeated} can be found in Section \ref{sec:pattern-rels}.

\begin{table}[H]
  \center
  \caption{Summary of Gothic headless relatives (repeated)}
    \input{tables/gothic-summary}
    \label{tbl:summary-gothic-repeated}
\end{table}

Old High German is another instance of a language in which the relative pronoun is allowed to surface in the internal case and the external case. This conclusion follows from my own research of the texts `Der althochdeutsche Isidor', `The Monsee fragments', `Otfrid's Evangelienbuch' and `Tatian' in ANNIS \citep{krause2016}.\footnote{
Old High German is widely discussed in the literature because of its case attraction in headed relatives \citep[cf.][]{pittner1995}, a phenomenon that seems related to case competition in headless relatives (see Section \ref{sec:attraction} for why attraction is not further discussed in this dissertation).
A common observation is that case attraction in headed relatives in Old High German adheres to the case scale. The same is claimed for headless relatives.
What, to my knowledge, has not been studied systematically is whether Old High German headless relatives allow the internal case and the external case to surface when either of them wins the case competition. This is what I investigated in my work.
}
The examples follow the spelling and the detailed glosses in ANNIS. The translations are my own.

First I discuss examples in which the internal and the external case match, and then examples in which they differ. If the internal case and the external case are identical, the relative pronoun simply surfaces in that case. I illustrate this for the nominative, the accusative and the dative.

Consider the example in \ref{ex:ohg-nom-nom}, in which the internal nominative case competes against the external nominative case.
The internal case is nominative, as the predicate \tit{senten} `to send' takes nominative subjects.
The external case is nominative as well, as the predicate \tit{queman} `to come' also takes nominative subjects.
The relative pronoun \tit{dher} `\ac{rel}.\ac{sg}.\ac{m}.\ac{nom}' appears in the internal and external case: the nominative.

\exg. quham dher chisendit scolda uuerdhan\\
 come.\ac{pst}.3\ac{sg}\scsub{[nom]} \ac{rel}.\ac{sg}.\ac{m}.\ac{nom} send.\ac{pst}.\ac{ptcp}\scsub{[nom]} should.\ac{pst}.3\ac{sg} become.\ac{inf}\\
 `the one, who should have been sent, came' \flushfill{Old High German, \ac{isid} 35:5}\label{ex:ohg-nom-nom}

Consider the example in \ref{ex:ohg-acc-acc}, in which the internal accusative case competes against the external accusative case.
The internal case is accusative, as the predicate \tit{quedan} `to speak' takes accusative objects.
The external case is accusative as well, as the predicate \tit{gihoren} `to listen to' also takes accusative objects.
The relative pronoun \tit{thiu} `\ac{rel}.\ac{pl}.\ac{n}.\ac{acc}' appears in the internal and external case: the accusative.

\exg. gihortut ir thiu ih íu quad\\
 listen.\ac{pst}.2\ac{pl}\scsub{[acc]} 2\ac{pl}.\ac{nom} \ac{rel}.\ac{pl}.\ac{n}.\ac{nom} 1\ac{sg}.\ac{nom} 2\ac{pl}.\ac{dat} speak.\ac{pst}.1\ac{sg}\scsub{[acc]}\\
 `you listened to those things, that I said to you' \flushfill{Old High German, \ac{tatian} 165:6}\label{ex:ohg-acc-acc}

Consider the example in \ref{ex:os-dat-dat}, in which the internal dative case competes against the external dative case.\footnote{
I could not find such an instance for this situation in any of the Old High German texts. This example comes from the `Heliand', an Old Saxon text written around the same time as the Old High German works I give examples from. Old Saxon is linguistically speaking the closest relative of Old High German.
}
The internal case is dative, as the predicate \tit{willian} `to wish' takes dative objects.
The external case is dative as well, as the predicate \tit{seggian} `to say' takes dative indirect objects.
The relative pronoun \tit{them} `\ac{rel}.\ac{pl}.\ac{m}.\ac{dat}' appears in the internal and external case: the dative.

\exg. sagda them siu uuelda\\
 say.\ac{pst}.3\ac{sg}\scsub{[dat]} \ac{rel}.\ac{pl}.\ac{m}.\ac{dat} 3\ac{sg}.\ac{f}.\ac{nom} wish.\ac{pst}.3\ac{sg}\scsub{[dat]}\\
 `she said to those, whom she wished for' \flushfill{Old Saxon, \ac{hel} 4:293}\label{ex:os-dat-dat}

These findings can be summarized as in Table \ref{tbl:summary-ohg-matching}. The top-left to bottom-right diagonal corresponds to the examples I have given so far in which the internal and external case match. The nominative marked in light gray corresponds to \ref{ex:ohg-nom-nom}, in which the internal nominative case competes against the external nominative case, and the relative pronoun surfaces in the nominative case. The accusative marked in dark gray corresponds to \ref{ex:ohg-acc-acc}, in which the internal accusative case competes against the external accusative case, and the relative pronoun surfaces in the accusative case. The unmarked dative corresponds to \ref{ex:os-dat-dat}, in which the internal dative case competes against the external dative case, and the relative pronoun surfaces in the dative case.

\begin{table}[H]
  \center
  \caption{Old High German headless relatives (matching)}
  \begin{tabular}{c|c|c|c}
    \toprule
     \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \colorbox{LG}{\ac{nom}}
          &
          &
          \\ \cmidrule{1-4}
      [\ac{acc}]
          &
          & \colorbox{DG}{\ac{acc}}
          &
          \\ \cmidrule{1-4}
      [\ac{dat}]
          &
          &
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
    \label{tbl:summary-ohg-matching}
\end{table}

In Table \ref{tbl:summary-ohg-matching}, six cells remain empty. These are the cases in which the internal and the external case differ. In the remainder of this section, I discuss them one by one.

I start with the competition between the accusative and the nominative. Following the case scale, the relative pronoun appears in the accusative case and never in nominative. As Old High German allows the internal and external case to surface, the accusative surfaces when it is the internal case and when it is the external case.

Consider the example in \ref{ex:ohg-nom-acc}. In this example, the internal accusative case competes against the external nominative case.
The internal case is accusative, as the predicate \tit{zellen} `to tell' takes accusative objects.
The external case is nominative, as the predicate \tit{sin} `to be' takes nominative objects.
The relative pronoun \tit{then} `\ac{rel}.\ac{sg}.\ac{m}.\ac{acc}' appears in the internal case: the accusative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.
Examples in which the internal case is accusative, the external case is nominative and the relative pronoun appears in the nominative case are unattested.

\exg. Thíz ist \tbf{then} \tbf{sie} \tbf{zéllent}\\
\ac{dem}.\ac{sg}.\ac{n}.\ac{nom} be.\ac{pres}.3\ac{sg}\scsub{[nom]} \ac{rel}.\ac{sg}.\ac{m}.\ac{acc} 3\ac{pl}.\ac{m}.\ac{nom} tell.\ac{pres}.3\ac{pl}\scsub{[acc]}\\
`this is the one whom they talk about' \flushfill{Old High German, \ac{otfrid} III 16:50}\label{ex:ohg-nom-acc}

Consider the example in \ref{ex:ohg-acc-nom}. In this example, the internal nominative case competes against the external accusative case.
The internal case is nominative, as the predicate \tit{gisizzen} `to possess' takes nominative subjects.
The external case is accusative, as the predicate \tit{bibringan} `to create' takes accusative objects.
The relative pronoun \tit{dhen} `\ac{rel}.\ac{sg}.\ac{m}.\ac{acc}' appears in the external case: the accusative. The relative pronoun is not marked in bold, just as the main clause, showing that the relative pronoun patterns with the main clause.\footnote{
At the end of this section I discuss a counterexample to the case scale, in which the internal case is nominative, the external case is accusative, and the relative pronoun appears in the nominative case.
}

\exg. ih bibringu fona iacobes samin endi fona iuda dhen \tbf{mina} \tbf{berga} \tbf{chisitzit}\\
1\ac{sg}.\ac{nom} {create}.\ac{pres}.1\ac{sg}\scsub{[acc]} of Jakob.\ac{gen} seed.\ac{sg}.\ac{dat} and of Judah.\ac{dat} \ac{rel}.\ac{sg}.\ac{m}.\ac{acc} my.\ac{acc}.\ac{m}.\ac{pl} mountain.\ac{acc}.\ac{pl} possess.\ac{pres}.3\ac{sg}\scsub{[nom]}\\
`I create of the seed of Jacob and of Judah the one, who possess my mountains' \flushfill{Old High German, \ac{isid} 34:3}\label{ex:ohg-acc-nom}

The two examples in which the nominative and the accusative compete are highlighted in Table \ref{tbl:summary-old-high-german-nom-acc}. The light gray marking corresponds to \ref{ex:ohg-nom-acc}, in which the internal accusative wins over the external nominative, and the relative pronoun surfaces in the accusative case. The dark gray marking corresponds to \ref{ex:ohg-acc-nom}, in which the external accusative wins over the internal nominative, and the relative pronoun surfaces in the accusative case.

\begin{table}[H]
  \center
  \caption{Old High German headless relatives (\ac{nom} --- \ac{acc})}
  \begin{tabular}{c|c|c|c}
    \toprule
        \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \cellcolor{DG}\ac{acc}
          &
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \cellcolor{LG}\ac{acc}
          & \ac{acc}
          &
          \\ \cmidrule{1-4}
      [\ac{dat}]
          &
          &
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
    \label{tbl:summary-old-high-german-nom-acc}
\end{table}

I continue with the competition between the dative and the nominative. Following the case scale, the relative pronoun appears in the dative case and never in nominative. As Old High German allows the internal and the external case to surface, the dative surfaces when it is the internal case and when it is the external case.

Consider the example in \ref{ex:ohg-nom-dat}. In this example, the internal dative case competes against the external nominative case.
The internal case is dative, as the predicate \tit{forlazan} `to read' takes dative indirect objects.
The external case is nominative, as the predicate \tit{minnon} `to love' takes nominative subjects.
The relative pronoun \tit{themo} `\ac{rel}.\ac{sg}.\ac{m}.\ac{dat}' appears in the internal case: the dative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.
Examples in which the internal case is dative, the external case is nominative and the relative pronoun appears in the nominative case are unattested.

\exg. \tbf{themo} \tbf{min} \tbf{uuirdit} \tbf{forlazan}, min minnot\\
\ac{rel}.\ac{sg}.\ac{m}.\ac{dat} less become.\ac{pres}.3\ac{sg} read.\ac{inf}\scsub{[dat]} less love.\ac{pres}.3\ac{sg}\scsub{[nom]}\\
`to whom less is read, loves less' \flushfill{Old High German, \ac{tatian} 138:13}\label{ex:ohg-nom-dat}

Consider the example in \ref{ex:ohg-dat-nom}. In this example, the internal nominative case competes against the external dative case.
The internal case is nominative, as the predicate \tit{sprehhan} `to speak' takes nominative subjects.
The external case is dative, as the predicate \tit{antwurten} `to reply' takes dative objects.
The relative pronoun \tit{demo} `\ac{rel}.\ac{sg}.\ac{m}.\ac{dat}' appears in the external case: the dative. The relative pronoun is not marked in bold, just as the main clause, showing that the relative pronoun patterns with the main clause.
Examples in which the internal case is nominative, the external case is dative and the relative pronoun appears in the nominative case are unattested.

\exg. enti aer {ant uurta} demo \tbf{zaimo} \tbf{sprah}\\
and 3\ac{sg}.\ac{m}.\ac{nom} reply.\ac{pst}.3\ac{sg}\scsub{[dat]} \ac{rel}.\ac{sg}.\ac{m}.\ac{dat} {to 3\ac{sg}.\ac{m}.\ac{dat}} speak.\ac{pst}.3\ac{sg}\scsub{[nom]}\\
`and he replied to the one who spoke to him' \flushfill{Old High German, \ac{mons} 7:24, adapted from \pgcitealt{pittner1995}{199}}\label{ex:ohg-dat-nom}

The two examples in which the nominative and the dative compete are highlighted in Table \ref{tbl:summary-old-high-german-nom-dat}. The light gray marking corresponds to \ref{ex:ohg-nom-dat}, in which the internal dative wins over the external nominative, and the relative pronoun surfaces in the dative case. The dark gray marking corresponds to \ref{ex:ohg-dat-nom}, in which the external dative wins over the internal nominative, and the relative pronoun surfaces in the dative case.

\begin{table}[H]
  \center
  \caption{Old High German headless relatives (\ac{nom} --- \ac{dat})}
  \begin{tabular}{c|c|c|c}
    \toprule
        \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{acc}
          & \cellcolor{DG}\ac{dat}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{acc}
          & \ac{acc}
          &
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \cellcolor{LG}\ac{dat}
          &
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
    \label{tbl:summary-old-high-german-nom-dat}
\end{table}

I end with the competition between the dative and the accusative. Following the case scale, the relative pronoun appears in the dative case and never in accusative. As Old High German allows the internal and the external case to surface, the dative surfaces when it is the internal case and when it is the external case.

Consider the example in \ref{ex:ohg-acc-dat}. In this example, the internal dative case competes against the external accusative case.
The internal case is dative, as the predicate \tit{zawen} `to tell' takes dative subjects.
The external case is accusative, as the predicate \tit{weizan} `to know' takes accusative objects.
The relative pronoun \tit{thémo} `\ac{rel}.\ac{sg}.\ac{m}.\ac{dat}' appears in the external case: the dative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.
Examples in which the internal case is accusative, the external case is dative and the relative pronoun appears in the accusative case are unattested.

\exg. weiz \tbf{thémo} \tbf{ouh} \tbf{baz} \tbf{záweta}\\
know.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{sg}.\ac{m}.\ac{dat} also better manage.\ac{pst}.3\ac{sg}\scsub{[dat]}\\
`I know the one who also managed it better' \flushfill{Old High German, \ac{otfrid} V 5:5}\label{ex:ohg-acc-dat}

Consider the example in \ref{ex:ohg-dat-acc}. In this example, the internal accusative case competes against the external dative case.
The internal case is accusative, as the predicate \tit{zellen} `to tell' takes accusative objects.
The external case is dative, as the comparative of the adjective \tit{furiro} `great' takes dative objects.
The relative pronoun \tit{thên} `\ac{rel}.\ac{pl}.\ac{m}.\ac{dat}' appears in the external case: the dative. The relative pronoun is not marked in bold, just as the main clause, showing that the relative pronoun patterns with the main clause.
Examples in which the internal case is accusative, the external case is dative and the relative pronoun appears in the accusative case are unattested.

\exg. bis -tú nu {zi wáre} furira Ábrahame? ouh thén \tbf{man} \tbf{hiar} \tbf{nu} \tbf{zálta}\\
be.\ac{pres}.2\ac{sg} -2\ac{sg}.\ac{nom} now truly {great}.\ac{cmpr}\scsub{[dat]} Abraham.\ac{dat} and \ac{rel}.\ac{pl}.\ac{m}.\ac{dat} one.\ac{nom}.\ac{m}.\ac{sg} here now tell.\ac{pst}.3\ac{sg}\scsub{[acc]}\\
`are you now truly greater than Abraham? and than those, who one talked about here now' \flushfill{Old High German, \ac{otfrid} III 18:33}\label{ex:ohg-dat-acc}

The two examples in which the accusative and the dative compete are highlighted in Table \ref{tbl:summary-old-high-german-acc-dat}. The light gray marking corresponds to \ref{ex:ohg-acc-dat}, in which the internal dative wins over the external accusative, and the relative pronoun surfaces in the dative case. The dark gray marking corresponds to \ref{ex:ohg-dat-acc}, in which the external dative wins over the internal accusative, and the relative pronoun surfaces in the dative case.

\begin{table}[H]
  \center
  \caption{Old High German headless relatives (\ac{acc} --- \ac{dat})}
  \begin{tabular}{c|c|c|c}
    \toprule
        \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{acc}
          & \ac{acc}
          & \cellcolor{DG}\ac{dat}
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \ac{dat}
          & \cellcolor{LG}\ac{dat}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
    \label{tbl:summary-old-high-german-acc-dat}
\end{table}

In my research I encountered a single counterexample to the pattern I just described.
Consider the example in \ref{ex:ohg-counterexample}. In this example, the internal nominative case competes against the external accusative case.
The internal case is nominative, as the predicate \tit{giheilen} `to save' takes nominative subjects.
The external case is accusative, as the predicate \tit{beran} `to bear' takes accusative objects.
Surprisingly, the relative pronoun \tit{thér} `\ac{rel}.\ac{sg}.\ac{m}.\ac{nom}' appears in the internal case: the nominative, which is the less complex of the two cases. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.

\exg. Tház si uns béran scolti \tbf{thér} \tbf{unsih} \tbf{gihéilti}\\
 that 3\ac{sg}.\ac{f}.\ac{nom} 1\ac{pl}.\ac{dat} bear.\ac{inf}\scsub{[acc]} should.\ac{subj}.\ac{pst}.3\ac{sg} \ac{rel}.\ac{sg}.\ac{m}.\ac{nom} 1\ac{pl}.\ac{acc} save.\ac{sbjv}.\ac{pst}.3\ac{sg}\scsub{[nom]}\\
 `that she should have beared for us the one, who had saved us' \flushfill{Old High German, \ac{otfrid} I 3:38}\label{ex:ohg-counterexample}

This example is unexpected, because the least complex case (the nominative) wins and not the most complex case (the accusative).
The only explanation for this I can see is a functional one. The \tit{thér} `\ac{rel}.\ac{sg}.\ac{m}.\ac{nom}' in \ref{ex:ohg-counterexample} refers to Jesus. In the relative clause he is the subject of \tit{unsih gihéilti} `had saved us', hence the internal nominative case. In the main clause he is the object of \tit{tház si uns béran scolti} `that she should have beared', hence the external accusative case.
Letting the relative pronoun surface in the internal case could be interpreted as emphasizing the role of Jesus as a savior, rather than him being the object of being given birth to. In line with that reasoning, it is expected that certain grammatical facts more often deviate from regular patterns if Jesus is involved. I leave investigating this prediction for future research.
Of course, this does not answer the question of what happens to the accusative case required by the external predicate. It also does not explain why not another emphasizing strategy is used, for instance forming a light-headed relative, which would leave space for two cases.
I acknowledge this example as a counterexample to the pattern I describe, but I do not change my generalization, as this is a single occurrence.

Leaving the counterexample aside, I conclude that Gothic and Old High German are both instances of languages that allow the internal and the external case to surface. The relative pronoun surfaces in the case that wins the case competition.


\section{Only internal case allowed}\label{sec:pattern-ii}

This section discusses the situation in which only the internal case is allowed to surface when it wins the case competition. When the internal case wins the case competition, the result is ungrammatical. I repeat the pattern from Section \ref{sec:possible-patterns} in Table \ref{tbl:case-competition-only-int-repeated}.

\begin{table}[H]
  \center
  \caption{Only internal case allowed (repeated)}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & *
           & *
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{acc}
           & \ac{acc}
           & *
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \ac{dat}
           & \ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-only-int-repeated}
\end{table}

An example of a language that shows this pattern is Modern German. In this section I discuss the Modern German data, based on the research of \citet{vogel2001}. The examples and the judgements are \posscitet{vogel2001}. I made the glosses more detailed, and I added translations where they were absent.

First I discuss examples in which the internal and the external case match, and then examples in which they differ. If the internal case and the external case are identical, the relative pronoun simply surfaces in that case. I illustrate this for the nominative, the accusative and the dative.

Consider the example in \ref{ex:mg-nom-nom}, in which the internal nominative case competes against the external nominative case.
The internal case is nominative, as the predicate \tit{mögen} `to like' takes nominative subjects.
The external case is nominative as well, as the predicate \tit{besuchen} `to visit' also takes nominative subjects.
The relative pronoun \tit{wer} `\ac{rel}.\ac{an}.\ac{nom}' appears in the internal and external case: the nominative.

\exg. Uns besucht, wer Maria mag.\\
 2\ac{pl}.\ac{acc} visit.\ac{pres}.3\ac{sg}\scsub{[nom]} \ac{rel}.\ac{an}.\ac{nom} Maria.\ac{acc} like.\ac{pres}.3\ac{sg}\scsub{[nom]}\\
 `Who visits us likes Maria.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{343}}\label{ex:mg-nom-nom}

Consider the example in \ref{ex:mg-acc-acc}, in which the internal accusative case competes against the external accusative case.
The internal case is accusative, as the predicate \tit{mögen} `to like' takes accusative objects.
The external case is accusative as well, as the predicate \tit{einladen} `to invite' also takes accusative objects.
The relative pronoun \tit{wen} `\ac{rel}.\ac{an}.\ac{acc}' appears in the internal and external case: the accusative.

\exg. Ich {lade ein}, wen auch Maria mag.\\
 1\ac{sg}.\ac{nom} invite.\ac{pres}.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{an}.\ac{acc} Maria.\ac{nom} like.\ac{pres}.3\ac{sg}\scsub{[acc]}\\
 `I invite who Maria also likes.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{344}}\label{ex:mg-acc-acc}

Consider the examples in \ref{ex:mg-dat-dat}, in which the internal dative case competes against the external dative case.
The internal case is dative, as the predicate \tit{vertrauen} `to please' takes dative objects.
The external case is dative as well, as the predicate \tit{folgen} `to follow' also takes dative objects.
The relative pronoun \tit{wem} `\ac{rel}.\ac{an}.\ac{dat}' appears in the internal and external case: the dative.

\exg. Ich folge, wem immer ich vertraue.\\
 1\ac{sg}.\ac{nom} folge.\ac{pres}.1\ac{sg}\scsub{[dat]} \ac{rel}.\ac{an}.\ac{dat} ever 1\ac{sg}.\ac{nom} vertraue.\ac{pres}.3\ac{sg}\scsub{[dat]}.\\
 `I follow whoever I trust.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{342}}\label{ex:mg-dat-dat}

These findings can be summarized as in Table \ref{tbl:summary-mg-matching}. The top-left to bottom-right diagonal corresponds to the examples I have given so far in which the internal and external case match. The nominative marked in light gray corresponds to \ref{ex:mg-nom-nom}, in which the internal nominative case competes against the external nominative case, and the relative pronoun surfaces in the nominative case. The accusative marked in dark gray corresponds to \ref{ex:mg-acc-acc}, in which the internal accusative case competes against the external accusative case, and the relative pronoun surfaces in the accusative case. The unmarked dative corresponds to \ref{ex:mg-dat-dat}, in which the internal dative case competes against the external dative case, and the relative pronoun surfaces in the dative case.

\begin{table}[H]
 \center
 \caption{Modern German headless relatives (matching)}
 \begin{tabular}{c|c|c|c}
   \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
         & [\ac{nom}]
         & [\ac{acc}]
         & [\ac{dat}]
         \\ \cmidrule{1-4}
     [\ac{nom}]
         & \colorbox{LG}{\ac{nom}}
         &
         &
         \\ \cmidrule{1-4}
     [\ac{acc}]
         &
         & \colorbox{DG}{\ac{acc}}
         &
         \\ \cmidrule{1-4}
     [\ac{dat}]
         &
         &
         & \ac{dat}
         \\
   \bottomrule
 \end{tabular}
   \label{tbl:summary-mg-matching}
\end{table}

In Table \ref{tbl:summary-mg-matching}, six cells remain empty. These are the cases in which the internal and the external case differ. In the remainder of this section, I discuss them one by one.

I start with the competition between the accusative and the nominative. Following the case scale, the relative pronoun appears in the accusative case and never in nominative. Following the internal-only requirement, when the accusative case is the internal case, the sentence is grammatical. When the accusative is the external case, the sentence is ungrammatical.

I start with the situation in which the internal case wins the competition, and it is possible to have a grammatical Modern German headless relative.
Consider the example in \ref{ex:mg-nom-acc}. In this example, the internal accusative case competes against the external nominative case.
The internal case is accusative, as the predicate \tit{mögen} `to like' takes accusative objects.
The external case is nominative, as the predicate \tit{besuchen} `to visit' takes nominative subjects.
The relative pronoun \tit{wen} `\ac{rel}.\ac{an}.\ac{acc}' appears in the internal case: the accusative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.
The example is grammatical, because the example adheres to the case scale, and the most complex case (here the accusative) is the internal case.

\exg. Uns besucht, \tbf{wen} \tbf{Maria} \tbf{mag}.\\
 2\ac{pl}.\ac{acc} visit.\ac{pres}.3\ac{sg}\scsub{[nom]} \ac{rel}.\ac{an}.\ac{acc} Maria.\ac{nom} like.\ac{pres}.3\ac{sg}\scsub{[acc]}\\
 `Who visits us, Maria likes.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{343}}\label{ex:mg-nom-acc}

The example in \ref{ex:mg-nom-acc-u} is identical to \ref{ex:mg-nom-acc}, except for that the relative pronoun appears in the external less complex nominative case. This example is ungrammatical: although the internal case is more complex, the relative pronoun appears in the least complex case (the nominative) and not in the most complex case (the accusative).

\exg. *Uns besucht, \tbf{wer} \tbf{Maria} \tbf{mag}.\\
 2\ac{pl}.\ac{acc} visit.\ac{pres}.3\ac{sg}\scsub{[nom]} \ac{rel}.\ac{an}.\ac{nom} Maria.\ac{nom} like.\ac{pres}.3\ac{sg}\scsub{[acc]}\\
 `Who visits us, Maria likes.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{343}}\label{ex:mg-nom-acc-u}

Now I turn to the situation in which the external case wins the competition, and there is no grammatical outcome possible, whichever case the relative pronoun appears in.
Consider the example in \ref{ex:mg-acc-nom}. In this example, the internal nominative case competes against the external accusative case.
The internal case is nominative, as the predicate \tit{sein} `to be' takes nominative subjects.
The external case is accusative, as the predicate \tit{einladen} `to invite' takes accusative objects.
The relative pronoun \tit{wen} `\ac{rel}.\ac{an}.\ac{acc}' appears in the external case: the accusative. The relative pronoun is not marked in bold, just as the main clause, showing that the relative pronoun patterns with the main clause.
The example adheres to the case scale, but the most complex case (here the accusative) is not the internal case. The example is ungrammatical, because only the internal can win the case competition in Modern German.

\exg. *Ich {lade ein}, wen \tbf{mir} \tbf{sympathisch} \tbf{ist}.\\
1\ac{sg}.\ac{nom} invite.\ac{pres}.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{an}.\ac{acc} 1\ac{sg}.\ac{dat} nice be.\ac{pres}.3\ac{sg}\scsub{[nom]}\\
`I invite who I like.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{344}}\label{ex:mg-acc-nom}

The example in \ref{ex:mg-acc-nom-u} is identical to \ref{ex:mg-acc-nom}, except for that the relative pronoun appears in the external less complex nominative case. This example is also ungrammatical: in addition to the most complex case not being the internal case, the relative pronoun also does not appear in the most complex case (the accusative) but in the least complex case (the nominative).\footnote{
Not every speaker or Modern German agrees with the ungrammaticality of \ref{ex:mg-acc-nom-u}. A sentence for which also has been claimed that speakers accept it is given in \ref{ex:mg-liebe-hasse}. This example was originally marked as ungrammatical by \pgcitet{groos1981}{206}.

\exg. Ich liebe wer gutes tut, und hasse, wer mich verletzt.\\
1\ac{sg}.\ac{nom} love.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{an}.\ac{nom} good.\ac{nmlz} do.3\ac{sg}\scsub{[nom]}
and hate.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{an}.\ac{nom} I\ac{sg}.\ac{acc} hurt.3\ac{sg}\scsub{[nom]}\\
`I love who does good and hate who hurts me.' \flushfill{Modern German, adapted from \pgcitealt{groos1981}{206}}\label{ex:mg-liebe-hasse}

The relative acceptability of \ref{ex:mg-acc-nom-u} and \ref{ex:mg-liebe-hasse} is unexpected because the relative pronoun appears in the least complex case (the nominative) and not in the more complex case (the accusative). However, the more complex case would also not be grammatical, because it is the external case, and Modern German only allows the relative pronoun to surface in the internal case. My hypothesis is that, because there is no way of making the headless relative grammatical, speakers try to make the construction work by somehow repairing it. I can think of two strategies for that: (1) they can take \tit{wer gutes tut} `who does good' and \tit{wer mich verletzt} `who hurts me' as clauses objects, which are not case-marked in German, or (2) they insert a morphologically silent object as the head of the relative clause.

Notice that this type of example is crucially different from the Old High German counterexample in \ref{ex:ohg-counterexample}. In the Old High German situation, there was a grammatical possibility which was not used, and in the Modern German situation, there is no grammatical way to make a headless relative.
}

\exg. *Ich {lade ein}, wer \tbf{mir} \tbf{sympathisch} \tbf{ist}.\\
1\ac{sg}.\ac{nom} invite.\ac{pres}.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{an}.\ac{nom} 1\ac{sg}.\ac{dat} nice be.\ac{pres}.3\ac{sg}\scsub{[nom]}\\
`I invite who I like.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{344}}\label{ex:mg-acc-nom-u}

The two examples in which the nominative and the accusative compete are highlighted in Table \ref{tbl:case-competition-mg-nom-acc}.
The light gray marking corresponds to \ref{ex:mg-nom-acc}, in which the internal accusative wins over the external nominative, and the relative pronoun surfaces in the accusative case (and not in the losing nominative case as in \ref{ex:mg-nom-acc-u}).
The dark gray marking corresponds to \ref{ex:mg-acc-nom}, in which the external accusative wins over the internal nominative, but the relative pronoun is not allowed to surface in the accusative case (or in the losing nominative case as in \ref{ex:mg-acc-nom-u}).

 \begin{table}[H]
   \center
   \caption{Modern German headless relatives (\ac{nom} --- \ac{acc})}
   \begin{tabular}{c|c|c|c}
     \toprule
     \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
            & [\ac{nom}]
            & [\ac{acc}]
            & [\ac{dat}]
            \\ \cmidrule{1-4}
        [\ac{nom}]
            & \ac{nom}
            & \cellcolor{DG}*
            &
            \\ \cmidrule{1-4}
        [\ac{acc}]
            & \cellcolor{LG}\ac{acc}
            & \ac{acc}
            &
            \\ \cmidrule{1-4}
        [\ac{dat}]
            &
            &
            & \ac{dat}
            \\
      \bottomrule
   \end{tabular}
     \label{tbl:case-competition-mg-nom-acc}
 \end{table}

I continue with the competition between the dative and the nominative. Following the case scale, the relative pronoun appears in the dative case and never in nominative. Following the internal-only requirement, when the dative case is the internal case, the sentence is grammatical.

I start again with the situation in which the internal case wins the competition, and it is possible to have a grammatical Modern German headless relative.
Consider the example in \ref{ex:mg-nom-dat}. In this example, the internal dative case competes against the external nominative case.
The internal case is dative, as the predicate \tit{vertrauen} `to trust' takes dative objects.
The external case is nominative, as the predicate \tit{besuchen} `to visit' takes nominative subjects.
The relative pronoun \tit{wem} `\ac{rel}.\ac{an}.\ac{dat}' appears in the internal case: the dative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.
The example adheres to the case scale, and the most complex case (here the dative) is the internal case, so the example is grammatical.

\exg. Uns besucht, \tbf{wem} \tbf{Maria} \tbf{vertraut}.\\
2\ac{pl}.\ac{acc} visit.\ac{pres}.3\ac{sg}\scsub{[nom]} \ac{rel}.\ac{an}.\ac{dat} Maria.\ac{nom} trust.\ac{pres}.3\ac{sg}\scsub{[dat]}\\
`Who visits us, Maria trusts.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{343}}\label{ex:mg-nom-dat}

The example in \ref{ex:mg-nom-dat-u} is identical to \ref{ex:mg-nom-dat}, except for that the relative pronoun appears in the external less complex nominative case. This example is ungrammatical: although the internal case is more complex, the relative pronoun appears in the least complex case (the nominative) and not in the most complex case (the dative).

\exg. *Uns besucht, \tbf{wer} \tbf{Maria} \tbf{vertraut}.\\
2\ac{pl}.\ac{acc} visit.\ac{pres}.3\ac{sg}\scsub{[nom]} \ac{rel}.\ac{an}.\ac{nom} Maria.\ac{nom} trust.\ac{pres}.3\ac{sg}\scsub{[dat]}\\
`Who visits us, Maria trusts.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{343}}\label{ex:mg-nom-dat-u}

Now I turn again to the situation in which the external case wins the competition, and there is no grammatical outcome possible, whichever case the relative pronoun appears in.
Consider the example in \ref{ex:mg-dat-nom}. In this example, the internal nominative case competes against the external dative case.
The internal case is nominative, as the predicate \tit{mögen} `to like' takes nominative subjects.
The external case is dative, as the predicate \tit{vertrauen} `to trust' takes dative objects.
The relative pronoun \tit{wem} `\ac{rel}.\ac{an}.\ac{dat}' appears in the external case: the dative. The relative pronoun is not marked in bold, just as the main clause, showing that the relative pronoun patterns with the main clause.
The example adheres to the case scale, but the most complex case (here the dative) is not the internal case. The example is ungrammatical, because only the internal can win the case competition in Modern German.

\exg. *Ich vertraue, wem \tbf{Hitchcock} \tbf{mag}.\\
1\ac{sg}.\ac{nom} trust.\ac{pres}.1\ac{sg}\scsub{[dat]} \ac{rel}.\ac{an}.\ac{dat} Hitchcock.\ac{acc} like.\ac{pres}.3\ac{sg}\scsub{[nom]}\\
`I trust who likes Hitchcock.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{345}}\label{ex:mg-dat-nom}

The example in \ref{ex:mg-dat-nom-u} is identical to \ref{ex:mg-dat-nom}, except for that the relative pronoun appears in the external less complex nominative case. This example is also ungrammatical: in addition to the most complex case not being the internal case, the relative pronoun also does not appear in the most complex case (the dative) but in the least complex case (the nominative).

\exg. *Ich vertraue, wer \tbf{Hitchcock} \tbf{mag}.\\
1\ac{sg}.\ac{nom} trust.\ac{pres}.1\ac{sg}\scsub{[dat]} \ac{rel}.\ac{an}.\ac{nom} Hitchcock.\ac{acc} like.\ac{pres}.3\ac{sg}\scsub{[nom]}\\
`I trust who likes Hitchcock.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{345}}\label{ex:mg-dat-nom-u}

The two examples in which the nominative and the dative compete are highlighted in Table \ref{tbl:case-competition-mg-nom-dat}.
The light gray marking corresponds to \ref{ex:mg-nom-dat}, in which the internal dative wins over the external nominative, and the relative pronoun surfaces in the dative case (and not in the losing nominative case as in \ref{ex:mg-nom-dat-u}).
The dark gray marking corresponds to \ref{ex:mg-dat-nom}, in which the external dative wins over the internal nominative, but the relative pronoun is not allowed to surface in the dative case (or in the losing nominative case as in \ref{ex:mg-dat-nom-u}).

\begin{table}[H]
  \center
  \caption{Modern German headless relatives (\ac{nom} --- \ac{dat})}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & *
           & \cellcolor{DG}*
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{acc}
           & \ac{acc}
           &
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \cellcolor{LG}\ac{dat}
           &
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-mg-nom-dat}
\end{table}

I end with the competition between the dative and the accusative. Following the case scale, the relative pronoun appears in the dative case and never in accusative. Following the internal-only requirement, when the dative case is the internal case, the sentence is grammatical.

I start again with the situation in which the internal case wins the competition, and it is possible to have a grammatical Modern German headless relative.
Consider the example in \ref{ex:mg-acc-dat}. In this example, the internal dative case competes against the external accusative case.
The internal case is dative, as the predicate \tit{vertrauen} `to trust' takes dative objects.
The external case is accusative, as the predicate \tit{einladen} `to invite' takes accusative objects.
The relative pronoun \tit{wem} `\ac{rel}.\ac{an}.\ac{dat}' appears in the internal case: the dative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.
The example adheres to the case scale, and the most complex case (here the dative) is the internal case, so the example is grammatical.

\exg. Ich {lade ein}, \tbf{wem} \tbf{auch} \tbf{Maria} \tbf{vertraut}. \\
1\ac{sg}.\ac{nom} invite.\ac{pres}.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{an}.\ac{dat} also Maria.\ac{nom} trust.\ac{pres}.3\ac{sg}\scsub{[dat]}.\\
`I invite whoever Maria also trusts.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{344}}\label{ex:mg-acc-dat}

The example in \ref{ex:mg-acc-dat-u} is identical to \ref{ex:mg-acc-dat}, except for that the relative pronoun appears in the external less complex accusative case. This example is ungrammatical: although the internal case is more complex, the relative pronoun appears in the least complex case (the accusative) and not in the most complex case (the dative).

\exg. *Ich {lade ein}, \tbf{wen} \tbf{auch} \tbf{Maria} \tbf{vertraut}. \\
1\ac{sg}.\ac{nom} invite.\ac{pres}.1\ac{sg}\scsub{[acc]} \ac{rel}.\ac{an}.\ac{acc} also Maria.\ac{nom} trust.\ac{pres}.3\ac{sg}\scsub{[dat]}.\\
`I invite whoever Maria also trusts.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{344}}\label{ex:mg-acc-dat-u}

Now I turn again to the situation in which the external case wins the competition, and there is no grammatical outcome possible, whichever case the relative pronoun appears in.
Consider the example in \ref{ex:mg-dat-acc}. In this example, the internal accusative case competes against the external dative case.
The internal case is accusative, as the predicate \tit{mögen} `to like' takes accusative objects.
The external case is dative, as the predicate \tit{vertrauen} `to trust' takes dative objects.
The relative pronoun \tit{wem} `\ac{rel}.\ac{an}.\ac{dat}' appears in the external case: the dative. The relative pronoun is not marked in bold, just as the main clause, showing that the relative pronoun patterns with the main clause.
The example adheres to the case scale, but the most complex case (here the dative) is not the internal case. The example is ungrammatical, because only the internal can win the case competition in Modern German.

\exg. *Ich vertraue, wem \tbf{auch} \tbf{Maria} \tbf{mag}. \\
1\ac{sg}.\ac{nom} trust.\ac{pres}.1\ac{sg}\scsub{[dat]} \ac{rel}.\ac{an}.\ac{dat} also Maria.\ac{nom} like.\ac{pres}.3\ac{sg}\scsub{[acc]}.\\
`I trust whoever Maria also likes.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{345}}\label{ex:mg-dat-acc}

The example in \ref{ex:mg-dat-acc-u} is identical to \ref{ex:mg-dat-acc}, except for that the relative pronoun appears in the external less complex accusative case. This example is also ungrammatical: in addition to the most complex case not being the internal case, the relative pronoun also does not appear in the most complex case (the dative) but in the least complex case (the accusative).

\exg. *Ich vertraue, wen \tbf{auch} \tbf{Maria} \tbf{mag}. \\
1\ac{sg}.\ac{nom} trust.\ac{pres}.1\ac{sg}\scsub{[dat]} \ac{rel}.\ac{an}.\ac{acc} also Maria.\ac{nom} like.\ac{pres}.3\ac{sg}\scsub{[acc]}.\\
`I trust whoever Maria also likes.' \flushfill{Modern German, adapted from \pgcitealt{vogel2001}{345}}\label{ex:mg-dat-acc-u}

The two examples in which the nominative and the dative compete are highlighted in Table \ref{tbl:case-competition-mg-acc-dat}.
The light gray marking corresponds to \ref{ex:mg-acc-dat}, in which the internal dative wins over the external accusative, and the relative pronoun surfaces in the dative case (and not in the losing accusative case as in \ref{ex:mg-acc-dat-u}).
The dark gray marking corresponds to \ref{ex:mg-dat-acc}, in which the external dative wins over the internal nominative, but the relative pronoun is not allowed to surface in the dative case (or in the losing accusative case as in \ref{ex:mg-dat-acc-u}).

\begin{table}[H]
  \center
  \caption{Modern German headless relatives (\ac{acc} --- \ac{dat})}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & *
           & *
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{acc}
           & \ac{acc}
           & \cellcolor{DG}*
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \ac{dat}
           & \cellcolor{LG}\ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-mg-acc-dat}
\end{table}

In sum, Modern German is an instance of a language that only allows the internal case to surface. The relative pronoun surfaces in the most complex case, but only when this more complex case is the internal case.


\section{Only external case allowed}\label{sec:pattern-iii}

This section discusses the situation in which only the external case is allowed to surface when it wins the case competition. When the internal case wins the case competition, the result is ungrammatical. I repeat the pattern from Section \ref{sec:possible-patterns} in Table \ref{tbl:case-competition-only-ext-repeated}.

\begin{table}[H]
  \center
  \caption{Only external case allowed (repeated)}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & *
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & *
           & *
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-only-ext-repeated}
\end{table}

To my knowledge, this pattern is not attested in any natural language, whether extinct or alive. Classical Greek has been mentioned in the literature both as a language of the third type (c.f. \citealt[120]{cinqueforthcoming}, who actually classifies Gothic also as such) and as a language of the first type \citep[cf.][41]{grosu1987}. I show that the correct description of Classical Greek is the latter, and that it patterns with Gothic and Old High German.\footnote{
It does seem to be the case that examples in which the external case wins over the internal case are more frequent in Classical Greek than examples in which the internal case wins over the external case (see \citealt{kakarikos2014} for numerous examples of the former type).
In this dissertation I do not address the question of why certain constructions and configurations are more frequent than others. My goal is to set up a system that generates the grammatical patterns and excludes the ungrammatical or unattested patterns.
}
I start with an example in which a more complex external case wins the case competition over a less complex internal case, and the relative pronoun surfaces in the external case.

Consider the example in \ref{ex:ag-dat-acc}. In this example, the internal accusative case competes against the external dative case.
The internal case is accusative, as the predicate \tit{tíktō} `to give birth to' takes accusative objects.
The external case is dative, as the predicate \tit{ékhō} `to provide' takes dative indirect objects.
The relative pronoun \tit{hō̃ͅ} `\ac{rel}.\ac{sg}.\ac{m}.\ac{dat}' appears in the internal case: the dative. The relative pronoun is not marked in bold, unlike as the relative clause, showing that the relative pronoun patterns with the main clause.

\exg. pãn {tò tekòn} trophḕn ékhei hō̃ͅ \tbf{án} \tbf{tékēͅ}\\
any parent.\ac{sg}.\ac{nom} food.\ac{sg}.\ac{acc} provide.\ac{pres}.3\ac{sg}\scsub{[dat]} \ac{rel}.\ac{sg}.\ac{m}.\ac{dat} \ac{mod} {gives birth}.\ac{aor}.3\ac{sg}\scsub{[acc]}\\
`any parent provides food to what he would have given birth to' \flushfill{Classical Greek, \ac{pl.men} 237e, adapted from \pgcitealt{kakarikos2014}{292}}\label{ex:ag-dat-acc}

This example is compatible with the picture of Classical Greek only allowing the external case to surface when it wins the competition. I repeat Table \ref{tbl:case-competition-only-ext-repeated} from the beginning of this section as Table \ref{tbl:case-competition-ag-poss1}, and I mark the cell that corresponds to the example in \ref{ex:ag-dat-acc} in gray.

\begin{table}[H]
  \center
  \caption{Classical Greek headless relatives possibility 1}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & *
           & \ac{acc}
           & \cellcolor{LG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & *
           & *
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-ag-poss1}
\end{table}

However, the example in \ref{ex:ag-dat-acc} is not only compatible with the external-only pattern. Considering only the example I have given so far, it is still possible for Classical Classical Greek to be of the internal-and-external type. I repeat Table \ref{tbl:case-competition-int-ext-repeated} from Section \ref{sec:pattern-i} as Table \ref{tbl:case-competition-ag-poss2}, and I mark the cell that corresponds to the example in \ref{ex:ag-dat-acc} in gray.

\begin{table}[H]
  \center
  \caption{Classical Greek headless relatives possibility 2}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{acc}
           & \ac{acc}
           & \cellcolor{LG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \ac{dat}
           & \ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-ag-poss2}
\end{table}

What sets Table \ref{tbl:case-competition-ag-poss1} and Table \ref{tbl:case-competition-ag-poss2} apart is the bottom-left corner of the table. These are cases in which the internal case wins the case competition.
In Table \ref{tbl:case-competition-ag-poss1} these examples are not allowed to surface, and in Table \ref{tbl:case-competition-ag-poss2} they are.
In what follows, I give an example in which a more complex internal case wins over a less complex external case. This indicates that Classical Greek cannot be of the type shown in Table \ref{tbl:case-competition-ag-poss1}, but is has to be of the type shown in Table \ref{tbl:case-competition-ag-poss2}. In other words, it is not of the type that only allows the external case to surface when it wins the case competition.

Consider the example in \ref{ex:ag-nom-acc}. In this example, the internal accusative case competes against the external nominative case.
The internal case is accusative, as the predicate \tit{philéō} `to love' takes accusative objects.
The external case is nominative, as the predicate \tit{apothnḗiskō} `to die' takes nominative subjects.
The relative pronoun \tit{hòn} `\ac{rel}.\ac{sg}.\ac{m}.\ac{acc}' appears in the internal case: the accusative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.\footnote{
The sentence in \ref{ex:ag-nom-acc} can also be analyzed as a headed relative, in which the relative clause modifies the phonologically empty subject of \tit{apothnḗiskō} `to die'. Then, however, more needs to be said about how it is possible for a relative clause to modify a phonologically empty element.
}

\exg. \tbf{hòn} \tbf{hoi} \tbf{theoì} \tbf{philoũsin} apothnḗͅskei néos\\
\ac{rel}.\ac{sg}.\ac{m}.\ac{acc} the god.\ac{pl} love.3\ac{pl}\scsub{[acc]} die.3\ac{sg}\scsub{[nom]} young\\
`He, whom the gods love, dies young.' \flushfill{Classical Greek, \ac{men.dd}, 125}\label{ex:ag-nom-acc}

This example shows that Classical Greek is not an instance of the third possible pattern, in which only the external case is allowed to surface. Instead, as illustrated by Table \ref{tbl:case-competition-classical-greek}, the language allows the internal case (marked light gray) and the external case (marked dark gray) to surface when either of them wins the case competition.

\begin{table}[H]
  \center
  \caption{Summary of Classical Greek headless relatives}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \cellcolor{LG}\ac{acc}
           & \ac{acc}
           & \cellcolor{DG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \ac{dat}
           & \ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-classical-greek}
\end{table}

I do not discuss more examples from Classical Greek than I did until now. This does not change anything about the point I am making here: the only kind of system that is compatible with the examples given is the one in which the internal and the external case are allowed to surface when either of them wins the case competition. For more examples in which the external case wins, I refer the reader to \pgcitet{kakarikos2014}{292-294}. An example in which the external dative wins over the internal nominative can be found in \citet{noussia2015}. I am not aware of an example in which the internal dative wins over the external accusative.

To sum up, to my knowledge, there is no language in which only the external case is allowed to surface when it wins the case competition, and the internal case is not. Classical Greek patterns with Gothic and Old High German in that is allows the internal and the external case to surface.



\section{Internal and external case not allowed}\label{sec:pattern-iv}

This section discusses the situation in which neither the internal nor the external case is allowed to surface when either of them wins the case competition. When the internal case or the external is more complex, the result is ungrammatical. In other words, when the internal and the external case differ, there is no grammatical headless relative construction possible. Only when the internal and external case are identical, it is possible to make headless relative. I repeat the pattern from Section \ref{sec:possible-patterns} in Table \ref{tbl:case-competition-none-repeated}.

\begin{table}[H]
  \center
  \caption{Neither internal nor external allowed (repeated)}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & *
           & *
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & *
           & \ac{acc}
           & *
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & *
           & *
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-none-repeated}
\end{table}

An example of a language that shows this pattern is Polish. In this section I discuss the Polish data, based on the research of \citet{citko2013} after \citet{himmelreich2017}. I only go through the case competition between accusative and dative, as only this data is discussed. This does not change anything about the point I am making here: the only kind of system that is compatible with the examples given is the one in which neither the internal nor the external case is allowed to surface. I made the glosses more detailed, and I added translations where they were absent.

First I discuss examples in which the internal and the external case match, and then examples in which they differ. If the internal case and the external case are identical, the relative pronoun simply surfaces in that case. I illustrate this for the nominative, the accusative and the dative.

Consider the example in \ref{ex:polish-acc-acc}, in which the internal accusative case competes against the external accusative case.
The internal case and external case are accusative, as the predicate \tit{lubić} `to like' in both clauses takes accusative objects.
The relative pronoun \tit{kogo} `\tsc{rel}.\tsc{acc}.\tsc{an}.\tsc{sg}' appears in the internal and external case: the accusative.

\exg. Jan lubi kogo -kolkwiek Maria lubi.\\
 Jan like.\tsc{3sg}\scsub{[acc]} \tsc{rel}.\tsc{acc}.\tsc{an}.\tsc{sg} ever Maria like.\tsc{3sg}\scsub{[acc]}\\
 `Jan likes whoever Maria likes.' \flushfill{Polish, adapted from \citealt{citko2013} after \pgcitealt{himmelreich2017}{17}}\label{ex:polish-acc-acc}

Consider the example in \ref{ex:polish-dat-dat}, in which the internal dative case competes against the external dative case.
The internal case is dative, as the predicate \tit{ufać} `to trust' takes dative objects.
The external case is dative as well, as the predicate \tit{pomagać} `to help' also takes dative objects.
The relative pronoun \tit{them} `\ac{rel}.\ac{pl}.\ac{an}.\ac{dat}' appears in the internal and external case: the dative.

\exg. Jan pomaga komu -kolkwiek ufa.\\
 Jan help.\tsc{3sg}\scsub{[dat]} \tsc{rel}.\tsc{dat}.\tsc{an}.\tsc{sg} ever trust.\tsc{3sg}\scsub{[dat]}\\
 `Jan helps whomever he trusts.' \flushfill{Polish, adapted from \citealt{citko2013} after \pgcitealt{himmelreich2017}{17}}\label{ex:polish-dat-dat}

These findings can be summarized as in Table \ref{tbl:summary-polish-matching}. The top-left to bottom-right diagonal corresponds to the examples I have given so far in which the internal and external case match. The accusative marked in light gray corresponds to \ref{ex:polish-acc-acc}, in which the internal accusative case competes against the external accusative case, and the relative pronoun surfaces in the accusative case. The dative marked in dark gray corresponds to \ref{ex:polish-dat-dat}, in which the internal dative case competes against the external dative case, and the relative pronoun surfaces in the dative case.

\begin{table}[H]
 \center
 \caption{Polish headless relatives (matching)}
 \begin{tabular}{c|c|c}
   \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-3}
      [\ac{acc}]
          & \cellcolor{LG}\ac{acc}
          &
          \\ \cmidrule{1-3}
      [\ac{dat}]
          &
          & \cellcolor{DG}\ac{dat}
          \\
    \bottomrule
 \end{tabular}
   \label{tbl:summary-polish-matching}
\end{table}

In Table \ref{tbl:summary-polish-matching}, two cells remain empty. These are the cases in which the internal and the external case differ. In the remainder of this section, I discuss them one by one.

I give examples from the case competition between accusative and dative. According to the case scale, the dative would win over the accusative. However, as neither the internal nor the external case is allowed to surface when either of them wins the case competition, all examples are ungrammatical.

I start with the situation in which the internal case wins the competition, and there is no grammatical outcome possible, whichever case the relative pronoun appears in.
Consider the example in \ref{ex:mg-acc-dat}. In this example, the internal dative case competes against the external accusative case.
The internal case is dative, as the predicate \tit{dokuczać} `to tease' takes dative objects.
The external case is accusative, as the predicate \tit{lubić} `to like' takes accusative objects.
The relative pronoun \tit{komu} `\ac{rel}.\ac{an}.\ac{dat}' appears in the internal case: the dative. The relative pronoun is marked in bold, just as the relative clause, showing that the relative pronoun patterns with the relative clause.
The example adheres to the case scale, but the internal case is not allowed to surface when it wins the case competition. Therefore, the example is ungrammatical.

\exg. *Jan lubi \tbf{komu} \tbf{-kolkwiek} \tbf{dokucza}.\\
Jan like.\tsc{3sg}\scsub{[acc]} \tsc{rel}.\tsc{dat}.\tsc{an}.\tsc{sg} ever tease.\tsc{3sg}\scsub{[dat]}\\
`Jan likes whoever he teases.' \flushfill{Polish, adapted from \citealt{citko2013} after \pgcitealt{himmelreich2017}{17}}\label{ex:polish-acc-dat}

The example in \ref{ex:polish-acc-dat-u} is identical to \ref{ex:polish-acc-dat}, except for that the relative pronoun appears in the external less complex accusative case. This example is also ungrammatical: the external case is less complex, and the external case is not allowed to surface when it wins the case competition.

\exg. *Jan lubi kogo \tbf{-kolkwiek} \tbf{dokucza}.\\
Jan like.\tsc{3sg}\scsub{[acc]} \tsc{rel}.\tsc{acc}.\tsc{an}.\tsc{sg} ever tease.\tsc{3sg}\scsub{[dat]}\\
`Jan likes whoever he teases.' \flushfill{Polish, adapted from \citealt{citko2013} after \pgcitealt{himmelreich2017}{17}}\label{ex:polish-acc-dat-u}

Now I turn to the situation in which the external case wins the competition, and there is no grammatical outcome possible, whichever case the relative pronoun appears in.
Consider the example in \ref{ex:polish-dat-acc}. In this example, the internal accusative case competes against the external dative case.
The internal case is accusative, as the predicate \tit{wpuścić} `to let' takes accusative objects.
The external case is dative, as the predicate \tit{ufać} `to trust' takes dative objects.
The relative pronoun \tit{komu} `\ac{rel}.\ac{an}.\ac{dat}' appears in the external case: the dative. The relative pronoun is not marked in bold, just as the main clause, showing that the relative pronoun patterns with the main clause.
The example adheres to the case scale, but the external case is (as the internal case) not allowed to surface when it wins the case competition. Therefore, the example is ungrammatical.

\exg. *Jan ufa komu \tbf{-kolkwiek} \tbf{wpuścil} \tbf{do} \tbf{domu}.\\
Jan trust.\tsc{3sg}\scsub{[dat]} \tsc{rel}.\tsc{dat}.\tsc{an}.\tsc{sg} ever let.\tsc{3sg}\scsub{[acc]} to home\\
`Jan trusts whoever he let into the house.' \flushfill{Polish, adapted from \citealt{citko2013} after \pgcitealt{himmelreich2017}{17}}\label{ex:polish-dat-acc}

The example in \ref{ex:polish-dat-acc-u} is identical to \ref{ex:polish-dat-acc}, except for that the relative pronoun appears in the internal less complex accusative case. This example is also ungrammatical: the internal case is less complex, and the internal case is not allowed to surface when it wins the case competition.

\exg. *Jan ufa \tbf{kogo} \tbf{-kolkwiek} \tbf{wpuścil} \tbf{do} \tbf{domu}.\\
Jan trust.\tsc{3sg}\scsub{[dat]} \tsc{rel}.\tsc{acc}.\tsc{an}.\tsc{sg} ever let.\tsc{3sg}\scsub{[acc]} to home\\
`Jan trusts whoever he let into the house.' \flushfill{Polish, adapted from \citealt{citko2013} after \pgcitealt{himmelreich2017}{17}}\label{ex:polish-dat-acc-u}

The two examples in which the accusative and the dative compete are highlighted in Table \ref{tbl:summary-polish-acc-dat}.
The light gray marking corresponds to \ref{ex:polish-acc-dat}, in which the internal dative wins over the external accusative, but the relative pronoun is not allowed to surface in the dative case (or in the losing accusative case as in \ref{ex:polish-acc-dat-u}).
The dark gray marking corresponds to \ref{ex:polish-dat-acc}, in which the external dative wins over the internal accusative, but the relative pronoun is not allowed to surface in the dative case (or in the losing accusative case as in \ref{ex:polish-dat-acc-u}).

\begin{table}[H]
  \center
  \caption{Polish headless relatives (\ac{acc} --- \ac{dat})}
  \begin{tabular}{c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-3}
       [\ac{acc}]
           & \ac{acc}
           & \cellcolor{DG}{*}
           \\ \cmidrule{1-3}
       [\ac{dat}]
           & \cellcolor{LG}{*}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:summary-polish-acc-dat}
\end{table}

In sum, Polish is an instance of a language that allows neither the internal case nor the external case to surface. When the internal and the external case differ in Polish, there is no way to form a grammatical headless relative construction.

\section{Summary}\label{sec:summary-3-patterns}

In case competition in headless relatives two aspects play a role. The first one is which case wins the case competition. It is a crosslinguistically stable fact that this is determined by the case scale in \ref{ex:case-scale-two-patterns-sum}, repeated from Chapter \ref{ch:recurring}. A case more to the right on the scale wins over a case more to the left on the scale.

\ex. \ac{nom} < \ac{acc} < \ac{dat}\label{ex:case-scale-two-patterns-sum}

This generates the pattern shown in Table \ref{tbl:case-competition-table}. The dative wins over the accusative and the nominative, and the accusative wins over the nominative.

\begin{table}[H]
  \center
  \caption{Relative pronoun follows case competition}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \cellcolor{DG}\ac{acc}
           & \cellcolor{DG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \cellcolor{LG}\ac{acc}
           & \ac{acc}
           & \cellcolor{DG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \cellcolor{LG}\ac{dat}
           & \cellcolor{LG}\ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-table}
\end{table}

The second aspect is whether the internal and the external case are allowed to surface when either of them wins the case competition. This differs across languages. There are four logical possibilities: (1) the internal and the external case are allowed to surface when either of them wins, (2) only the internal case is allowed to surface when it wins, (3) only the external case is allowed to surface when it wins, and (4) neither the internal nor the external case is allowed to surface when either of them wins.

Gothic, Old High German and Classical Greek are examples of languages of the first type. In these languages, relative pronouns in the unmarked, light gray and dark gray cells in Table \ref{tbl:case-competition-table} are attested.
Modern German is an example of a language of the second type. In this language, relative pronouns in the unmarked and light gray cells in Table \ref{tbl:case-competition-table} are grammatical.
Polish is an example of a language of the fourth type. In this language, relative pronoun in only in the unmarked cells in Table \ref{tbl:case-competition-table} are grammatical.
To my knowledge, the third pattern is not attested. This would be a language in which relative pronoun are grammatical in the unmarked and the dark gray cells in Table \ref{tbl:case-competition-table}.

The crosslinguistically attested patterns can be generated by introducing two parameters, which is shown in Figure \ref{fig:attested-headless-relatives-case-competition}.
The first parameter is whether a language allows the internal case to surface or whether it does not at `\ac{int} as winner?'.
If a language does not allow the internal case to surface, the fourth pattern is generated. In this pattern, neither the internal nor the external case is allowed to surface when either of them wins, and only the the unmarked cells in Table \ref{tbl:case-competition-table} are grammatical.
If a language allows the internal case to surface, the second parameter is whether a language allows the external case to surface or whether it does not at `\ac{ext} as winner?'.
If a language does not allow the external case to surface, the second pattern is generated. In this pattern, only the internal case is allowed to surface when it wins, and the unmarked and light gray cells in Table \ref{tbl:case-competition-table} are grammatical.
If a language allows the external case to surface, the first pattern is generated. In this pattern, the internal and the external case are allowed to surface when either of them wins, and the unmarked light gray and the dark gray cells in Table \ref{tbl:case-competition-table} are grammatical.

\begin{figure}[H]
  \centering
    \footnotesize{
    \begin{tikzpicture}[node distance=1.5cm]
      \node (question2) [question]
      {\ac{int} as winner?};
          \node (outcome2) [outcome, below of=question2, xshift=-1.5cm]
          {*};
              \node (example2) [example, below of=outcome2, yshift=0.25cm]
              {\scriptsize{e.g. Polish\\\phantom{x}}};
          \node (question3) [question, below of=question2, xshift=2cm, yshift=-0.5cm]
          {\ac{ext} as winner?};
              \node (outcome3) [outcome, below of=question3, xshift=-1.5cm]
              {\ac{int} + complex};
                  \node (example3) [example, below of=outcome3, yshift=0.25cm]
                  {\scriptsize{e.g. Modern German\\\phantom{x}}};
              \node (outcome4) [outcome, below of=question3, xshift=1.5cm]
              {complex};
                  \node (example3) [example, below of=outcome4, yshift=0.25cm]
                  {\scriptsize{e.g. Gothic, Old High German, Classical Greek}};

    \draw [arrow] (question2) -- node[anchor=east] {no} (outcome2);
    \draw [arrow] (question2) -- node[anchor=west] {yes} (question3);
    \draw [arrow] (question3) -- node[anchor=east] {no} (outcome3);
    \draw [arrow] (question3) -- node[anchor=west] {yes} (outcome4);
    \end{tikzpicture}
    }
    \caption{Attested patterns in headless relatives with case competition}
    \label{fig:attested-headless-relatives-case-competition}
\end{figure}

These parameters necessarily follow each other in the order specified: `\ac{int} as winner?' precedes `\ac{ext} as winner?'. If they were ordered the other way around, the third pattern in which only the external case is allowed to surface when it wins, would be generated. Moreover, the second pattern, in which only the internal case is allowed to surface when it wins, would not be generated. The ordering in Figure \ref{fig:attested-headless-relatives-case-competition} ensures that the internal-only pattern is generated and the external-only is not, which is in line with what is attested crosslinguistically. In Chapter \ref{ch:relativization}, I discusses the linguistic counterpart of this ordering.


\section{Aside: languages without case competition}\label{sec:without-case-competition}

In this chapter so far, I discussed languages that show case competition. There are also languages that do not show any case competition. This section discusses these languages, and gives a typology of headless relatives.

In languages without case competition, the internal and external case do not compete to show their case on the relative pronoun. It is irrelevant how the two cases relate to each other on the case scale. Instead, it is fixed per language whether the relative pronoun appears in the internal or the external case. Logically, there are two possible languages without case competition: one that lets the relative pronoun appear in the internal case, and one that lets the relative pronoun appear in the external case.

Table \ref{tbl:no-case-competition-int} shows the pattern of a language in which the relative pronoun always appears in the internal case. In the second row, the internal case is nominative and the external case is nominative, accusative or dative. The relative pronoun appears in the nominative. It is irrelevant here that the nominative is less complex than the accusative and the dative, because there is no case competition taking place. The third row shows that the relative pronoun always appears in the accusative when the internal case is the accusative, and the fourth row shows the same for the dative. To my knowledge, this type is not attested in any natural language.

\begin{table}[H]
  \center
  \caption{Always internal case}
  \begin{tabular}{c|c|c|c}
    \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{nom}
          & \ac{nom}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{acc}
          & \ac{acc}
          & \ac{acc}
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \ac{dat}
          & \ac{dat}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
  \label{tbl:no-case-competition-int}
\end{table}

Table \ref{tbl:no-case-competition-ext} shows the pattern of a language in which the relative pronoun always appears in the external case. In the second column, the external case is nominative and the internal case is nominative, accusative or dative. The relative pronoun appears in the nominative. It is irrelevant here that the nominative is less complex than the accusative and the dative, because there is no case competition taking place. The third column shows that the relative pronoun always appears in the accusative when the external case is the accusative, and the fourth column shows the same for the dative.

\begin{table}[H]
  \center
  \caption{Always external case}
  \begin{tabular}{c|c|c|c}
    \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
  \label{tbl:no-case-competition-ext}
\end{table}

Section \ref{sec:always-ext} discusses two languages that let their relative pronouns in headless relatives always surface in the external case: Old English and Modern Greek. In Section \ref{sec:typology} I extend the typology from Section \ref{sec:summary-3-patterns} by adding the languages without case competition. As I briefly mentioned, I do not know of any language, whether extinct or alive, that lets the relative pronoun always surface in the internal case. I do not offer an explanation for why it is not attested, and I include this possibility in my typology.

\subsection{Always external case}\label{sec:always-ext}

In this section I discuss two languages in which the relative pronoun always appears in the external case. I show that these languages do not show any case competition. In other words, these languages are of the type shown in Table \ref{tbl:no-case-competition-ext} and not of the type I discussed in Section \ref{sec:pattern-iii} (or of the one in Section \ref{sec:pattern-i}).

\begin{table}[H]
  \center
  \caption{Always external case (repeated)}
  \begin{tabular}{c|c|c|c}
    \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
  \label{tbl:no-case-competition-ext-repeated}
\end{table}

Two example of languages that shows this pattern are Old English and Modern Greek. In this section I discuss the Old English data with examples from \citet{harbert1983}. The Modern Greek data I discuss is taken from \citet{daskalaki2011}. For all examples holds that I made the glosses more detailed, and I added and modified translations.

I start with Old English. I give an example in which the external case is more complex than the internal case and the relative pronoun appears in the most complex external case.

Consider the example in \ref{ex:oe-dat-nom}.
The internal case is nominative, as the predicate \tit{gegyltan} `to sin' takes nominative subjects.
The external case is dative, as the predicate \tit{for-gifan} `to forgive' takes dative objects.
The relative pronoun \tit{ðam} `\ac{rel}.\ac{dat}.\ac{pl}' appears in the external case: the dative. The relative pronoun is not marked in bold, unlike the relative clause, showing that the relative pronoun patterns with the main clause.

\exg. ðæt is, ðæt man for-gife, ðam \tbf{ðe} \tbf{wið} \tbf{hine} \tbf{gegylte}\\
 that is that one forgive.\ac{subj}.\ac{sg}\scsub{[dat]} \ac{rel}.\ac{dat}.\ac{pl} \ac{comp} against 3\ac{sg}.\ac{m}.\ac{acc} sin.3\ac{sg}\scsub{[nom]}\\
 `that is, that one₂ forgive him₁, who sins against him₂' \flushfill{Old English, adapted from \pgcitealt{harbert1983}{549}} \label{ex:oe-dat-nom}

This example is compatible with three patterns. First, Old English could be a case competition language that only allows the external case to surface. I repeat Table \ref{tbl:case-competition-only-ext-repeated} from Section \ref{sec:pattern-iii} as Table \ref{tbl:oe-poss1}, and I mark the cell that corresponds to example \ref{ex:oe-dat-nom} in gray.

 \begin{table}[H]
   \center
   \caption{Old English headless relatives possibility 1}
   \begin{tabular}{c|c|c|c}
     \toprule
     \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
            & [\ac{nom}]
            & [\ac{acc}]
            & [\ac{dat}]
            \\ \cmidrule{1-4}
        [\ac{nom}]
            & \ac{nom}
            & \ac{acc}
            & \cellcolor{LG}\ac{dat}
            \\ \cmidrule{1-4}
        [\ac{acc}]
            & *
            & \ac{acc}
            & \ac{dat}
            \\ \cmidrule{1-4}
        [\ac{dat}]
            & *
            & *
            & \ac{dat}
            \\
      \bottomrule
   \end{tabular}
     \label{tbl:oe-poss1}
 \end{table}

Second, Old English could be a case competition language that allows the internal case and the external case to surface. I repeat Table \ref{tbl:case-competition-int-ext-repeated} from Section \ref{sec:pattern-i} as Table \ref{tbl:oe-poss2}, and I mark the cell that corresponds to example \ref{ex:oe-dat-nom} in gray.

  \begin{table}[H]
    \center
    \caption{Old English headless relatives possibility 2}
    \begin{tabular}{c|c|c|c}
      \toprule
      \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
             & [\ac{nom}]
             & [\ac{acc}]
             & [\ac{dat}]
             \\ \cmidrule{1-4}
         [\ac{nom}]
             & \ac{nom}
             & \ac{acc}
             & \cellcolor{LG}\ac{dat}
             \\ \cmidrule{1-4}
         [\ac{acc}]
             & \ac{acc}
             & \ac{acc}
             & \ac{dat}
             \\ \cmidrule{1-4}
         [\ac{dat}]
             & \ac{dat}
             & \ac{dat}
             & \ac{dat}
             \\
       \bottomrule
    \end{tabular}
      \label{tbl:oe-poss2}
  \end{table}

Third, Old English could be a language without case competition that lets the relative pronoun appear in the external case. I repeat Table \ref{tbl:no-case-competition-ext-repeated} from the beginning of this section as Table \ref{tbl:oe-poss3}, and I mark the cell that corresponds to example \ref{ex:oe-dat-nom} in gray.

 \begin{table}[H]
   \center
   \caption{Old English headless relatives possibility 3}
   \begin{tabular}{c|c|c|c}
     \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \ac{acc}
           & \cellcolor{LG}\ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\
     \bottomrule
   \end{tabular}
   \label{tbl:oe-poss3}
 \end{table}

What sets Table \ref{tbl:oe-poss1}, Table \ref{tbl:oe-poss2} and Table \ref{tbl:oe-poss3} apart is the bottom-left corner of the table. These are situations in which the internal case is more complex than the external case.
In Table \ref{tbl:oe-poss1} the winning case is not allowed to surface, and there is no grammatical headless relative possible. If this is the pattern that Old English shows, then it would be a language with case competition that only allows the external case to surface, i.e. it would be of the type of Section \ref{sec:pattern-iii} I claimed did not exist.
In Table \ref{tbl:oe-poss2} and in Table \ref{tbl:oe-poss3} there is a relative pronoun that can surface, but the case of the relative pronouns differs. In Table \ref{tbl:oe-poss2}, the relative pronoun surfaces in the most complex case that wins the case competition: the internal case. In Table \ref{tbl:oe-poss3}, there is no case competition taking place, and the relative pronoun surfaces in the external case.

In the example that follows I show that Old English is of the type in Table \ref{tbl:oe-poss3}. I give an example in which the internal case is more complex than the external one. Nevertheless, the relative pronoun surfaces in the less complex external case. Old English is namely a language without case competition that lets the relative pronoun surface in the external case.

Consider the example in \ref{ex:oe-acc-dat}.
The internal case is dative, as the preposition \tit{onuppan} `upon' takes dative objects.
The external case is accusative, as the predicate \tit{tōbrȳsan} `to pulversize' takes accusative objects.
The relative pronoun \tit{ðone} `\ac{rel}.\ac{sg}.\ac{acc}' appears in the external case: the accusative.
The relative pronoun appears in the external case, although it is the least complex case of the two. The example is grammatical, because Old English does not show case competition, so the case scale is irrelevant. As long as the relative pronoun appears in the external case, the headless relative is grammatical.

\exg. he tobryst ðone \tbf{ðe} \tbf{he} \tbf{onuppan} \tbf{fylð}\\
 it pulverizes\scsub{[acc]} \ac{rel}.\ac{sg}.\ac{acc} \ac{comp} it upon\scsub{[dat]} falls\\
`It pulverizes him whom it falls upon.' \flushfill{Old English, adapted from \pgcitealt{harbert1983}{550}} \label{ex:oe-acc-dat}

This example shows that Old English is neither an instance of the pattern in Section \ref{sec:pattern-iii}, in which only the external case is allowed to surface, nor is it an instance of the pattern in Section \ref{sec:pattern-i}, in which the internal case and external case are allowed to surface.
Instead, as illustrated by Table \ref{tbl:no-case-competition-old-english}, the language does not have any case competition. The relative pronoun appears in the external case: the external case can be the most complex case, illustrated by the example in \ref{ex:oe-dat-nom}, marked here in light gray, or it can be the least complex case, illustrated by the example in \ref{ex:oe-acc-dat}, marked here in dark gray.

\begin{table}[H]
  \center
  \caption{Summary of Old English headless relatives}
  \begin{tabular}{c|c|c|c}
    \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{acc}
          & \cellcolor{LG}\ac{dat}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \ac{nom}
          & \cellcolor{DG}\ac{acc}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
  \label{tbl:no-case-competition-old-english}
\end{table}

I do not discuss more examples from Old English than I did until now. This does not change anything about the point I am making here: the only kind of system that is compatible with the examples given is the one in which the relative pronoun always appears in the external case.

The same pattern appears in Modern Greek. The only difference is that Modern Greek has the genitive, and not the dative. I start again with an example in which the external case is more complex than the internal case and the relative pronoun appears in the most complex external case.

Consider the example in \ref{ex:greek-acc-nom}.
The internal case is nominative, as the predicate \tit{voíθisó} `to help' takes nominative subjects.
The external case is accusative, as the predicate \tit{efχarístisó} `to thank' takes accusative objects.
The relative pronoun \tit{ópjus} `\ac{rel}.\ac{pl}.\ac{m}.\ac{acc}' appears in the external case: the accusative. The relative pronoun is not marked in bold, unlike the relative clause, showing that the relative pronoun patterns with the main clause.

\exg. Efχarístisa ópjus \tbf{me} \tbf{voíϑisan}.\\
thank.\ac{pst}.3\ac{pl}\scsub{[acc]} \ac{rel}.\ac{pl}.\ac{m}.\ac{acc} \ac{cl}.1\ac{sg}.\ac{acc} help.\ac{pst}.3\ac{pl}\scsub{[nom]}\\
`I thanked whoever helped me.' \flushfill{Modern Greek, adapted from \pgcitealt{daskalaki2011}{80}}\label{ex:greek-acc-nom}

This example is compatible with three patterns. First, Modern Greek could be a case competition language that only allows the external case to surface. I repeat Table \ref{tbl:case-competition-only-ext-repeated} from Section \ref{sec:pattern-iii} as Table \ref{tbl:greek-poss1}, and I mark the cell that corresponds to example \ref{ex:greek-acc-nom} in gray.

 \begin{table}[H]
   \center
   \caption{Modern Greek headless relatives possibility 1}
   \begin{tabular}{c|c|c|c}
     \toprule
     \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
            & [\ac{nom}]
            & [\ac{acc}]
            & [\ac{gen}]
            \\ \cmidrule{1-4}
        [\ac{nom}]
            & \ac{nom}
            & \cellcolor{LG}\ac{acc}
            & \ac{gen}
            \\ \cmidrule{1-4}
        [\ac{acc}]
            & *
            & \ac{acc}
            & \ac{gen}
            \\ \cmidrule{1-4}
        [\ac{gen}]
            & *
            & *
            & \ac{dat}
            \\
      \bottomrule
   \end{tabular}
     \label{tbl:greek-poss1}
 \end{table}

Second, Modern Greek could be a case competition language that allows the internal case and external case to surface. I repeat Table \ref{tbl:case-competition-int-ext-repeated} from Section \ref{sec:pattern-i} as Table \ref{tbl:greek-poss2}, and I mark the cell that corresponds to example \ref{ex:greek-acc-nom} in gray.

  \begin{table}[H]
    \center
    \caption{Modern Greek headless relatives possibility 2}
    \begin{tabular}{c|c|c|c}
      \toprule
      \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
             & [\ac{nom}]
             & [\ac{acc}]
             & [\ac{gen}]
             \\ \cmidrule{1-4}
         [\ac{nom}]
             & \ac{nom}
             & \cellcolor{LG}\ac{acc}
             & \ac{gen}
             \\ \cmidrule{1-4}
         [\ac{acc}]
             & \ac{acc}
             & \ac{acc}
             & \ac{gen}
             \\ \cmidrule{1-4}
         [\ac{gen}]
             & \ac{gen}
             & \ac{gen}
             & \ac{dat}
             \\
       \bottomrule
    \end{tabular}
      \label{tbl:greek-poss2}
  \end{table}

Third, Modern Greek could be a language without case competition that lets the relative pronoun appear in the external case. I repeat Table \ref{tbl:no-case-competition-ext-repeated} from the beginning of this section as Table \ref{tbl:greek-poss3}, and I mark the cell that corresponds to example \ref{ex:greek-acc-nom} in gray.

 \begin{table}[H]
   \center
   \caption{Modern Greek headless relatives possibility 3}
   \begin{tabular}{c|c|c|c}
     \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{gen}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \cellcolor{LG}\ac{acc}
           & \ac{gen}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{nom}
           & \ac{acc}
           & \ac{gen}
           \\ \cmidrule{1-4}
       [\ac{gen}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\
     \bottomrule
   \end{tabular}
   \label{tbl:greek-poss3}
 \end{table}

What sets Table \ref{tbl:greek-poss1}, Table \ref{tbl:greek-poss2} and Table \ref{tbl:greek-poss3} apart is the bottom-left corner of the table. These are cases in which the internal case is more complex than the external case.
In Table \ref{tbl:greek-poss1} the winning case is not allowed to surface, and there is no grammatical headless relative possible. If this is the pattern that Modern Greek shows, then it would be a language with case competition that only allows the external case to surface, i.e. it would be of the type of Section \ref{sec:pattern-iii} I claimed did not exist.
In Table \ref{tbl:greek-poss2} and in Table \ref{tbl:greek-poss3} there is a relative pronoun that can surface, but the case of the relative pronouns differs. In Table \ref{tbl:greek-poss2}, the relative pronoun surfaces in the most complex case that wins the case competition: the internal case. In Table \ref{tbl:greek-poss3}, there is no case competition taking place, and the relative pronoun surfaces in the external case.

In the example that follows I show that Modern Greek is of the type in Table \ref{tbl:greek-poss3}. I give an example in which the internal case is more complex than the external one. Nevertheless, the relative pronoun surfaces in the less complex external case. Modern Greek is namely a language without case competition that lets the relative pronoun surface in the external case.

Consider the example in \ref{ex:greek-nom-acc}.
The internal case is accusative, as the predicate \tit{irθó} `to invite' takes accusative objects.
The external case is nominative, as the predicate \tit{kálesó} `to come' takes nominative subjects.
The relative pronoun \tit{ópji} `\ac{rel}.\ac{pl}.\ac{m}.\ac{nom}' appears in the external case: the nominative.
The relative pronoun appears in the external case, although it is the least complex case of the two. The example is grammatical, because Modern Greek does not show case competition, so the case scale is irrelevant. As long as the relative pronoun appears in the external case, the headless relative is grammatical.

\exg. Irθan ópji \tbf{káleses}.\\
come.\ac{pst}.3\ac{pl}\scsub{[nom]} \ac{rel}.\ac{pl}.\ac{m}.\ac{nom} invite.\ac{pst}.2\ac{sg}\scsub{[acc]}\\
`Whoever you invited came.'\flushfill{Modern Greek, adapted from \pgcitealt{daskalaki2011}{80}}\label{ex:greek-nom-acc}

The example in \ref{ex:greek-nom-acc-u} is identical to \ref{ex:greek-nom-acc}, except for that the relative pronoun appears in the internal  more complex case. This example is ungrammatical: the relative pronoun does not appear in the external case. The fact that the internal case is more complex is irrelevant.

\exg. *Irθan \tbf{ópjus} \tbf{káleses}.\\
come.\ac{pst}.3\ac{pl}\scsub{[nom]} \ac{rel}.\ac{pl}.\ac{m}.\ac{acc} invite.\ac{pst}.2\ac{sg}\scsub{[acc]}\\
`Whoever you invited came.'\flushfill{Modern Greek, adapted from \pgcitealt{daskalaki2011}{79}}\label{ex:greek-nom-acc-u}

This example shows that Modern Greek is neither an instance of the pattern in Section \ref{sec:pattern-iii}, in which only the external case is allowed to surface, nor is it an instance of the pattern in Section \ref{sec:pattern-i}, in which the internal case and external case are allowed to surface.
Instead, as illustrated by Table \ref{tbl:no-case-competition-greek}, the language does not have any case competition. The relative pronoun appears in the external case: the external case can be the most complex case, illustrated by the example in \ref{ex:greek-acc-nom}, marked here in light gray, or it can be the least complex case, illustrated by the example in \ref{ex:greek-nom-acc}, marked here in dark gray.

\begin{table}[H]
  \center
  \caption{Summary of Modern Greek headless relatives}
  \begin{tabular}{c|c|c|c}
    \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{gen}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \cellcolor{LG}\ac{acc}
          & \ac{gen}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \cellcolor{DG}\ac{nom}
          & \ac{acc}
          & \ac{gen}
          \\ \cmidrule{1-4}
      [\ac{gen}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
  \label{tbl:no-case-competition-greek}
\end{table}

There is something more to be said about the situation in Modern Greek. When the internal case is genitive instead of accusative, a clitic is added to the sentence to make it grammatical.

Consider the example in \ref{ex:greek-nom-gen}.
The internal case is genitive, as the predicate \tit{eðósó} `to give' takes genitive objects.
The external case is accusative, as the predicate \tit{efχarístisó} `to thank' takes nominative subjects.
The relative pronoun \tit{ópjon} `\ac{rel}.\ac{pl}.\ac{m}.\ac{nom}' appears in the external case: the nominative.
The relative pronoun appears in the external case, although it is the least complex case of the two. The example is grammatical, because Modern Greek does not show case competition, so the case scale is irrelevant. As long as the relative pronoun appears in the external case, the headless relative is grammatical. In addition, the relative clause obligatorily contains the genitive clitic \tit{tus} `\ac{cl}.3\ac{pl}.\ac{gen}'.\footnote{
In Modern German, it is possible to insert a light head to resolve a situation with a more complex external case. However, then the relative pronoun has to change as well (from a \tsc{wh}-pronoun into a \tsc{d}-pronoun). I assume this is a different construction, and the Modern Greek one with the clitic inserted is not.
}

\exg. Me efχarístisan ópji \tbf{tus} \tbf{íχa} \tbf{ðósi} \tbf{leftá}.\\
 \ac{cl}.1\ac{sg}.\ac{acc} thank.\ac{pst}.3\ac{pl}\scsub{[nom]} \ac{rel}.\ac{pl}.\ac{m}.\ac{nom} \ac{cl}.3\ac{pl}.\ac{gen} have.\ac{pst}.1\ac{sg} give.\ac{ptcp}\scsub{[gen]} money\\
 `Whoever I had given money to, thanked me.'\flushfill{Modern Greek, adapted from \pgcitealt{daskalaki2011}{80}}\label{ex:greek-nom-gen}

This once again confirms the picture of Modern Greek always letting the relative pronoun surface in the external case. The internal case is taken care of by the clitic, which is independent of the relative clause construction.

I do not discuss more examples from Modern Greek than I did until now. This does not change anything about the point I am making here: the only kind of system that is compatible with the examples given is the one in the relative pronoun always appears in the external case. For more examples that illustrate this pattern, I refer the reader to \pgcitet{daskalaki2011}{79-80}.

In sum, Old English and Modern Greek are languages without case competition in their headless relatives. The relative pronoun always appears in the external case.

\subsection{A typology of headless relatives}\label{sec:typology}

This section provides a typological overview of headless relatives. First, I describe the difference between the patterns of languages with and without case competition. Second, I include the parameters of non-case competition languages in the diagram I introduced in Section \ref{sec:summary-3-patterns}. Third, I give an overview of all logically possible patterns, and I show how the diagram generates the attested ones and does not generate the ones that are not attested.

In Section \ref{sec:pattern-i} to \ref{sec:pattern-iv}, I discussed four different patterns that are all derived from the same table. The relative pronoun appears in the case that wins the case competition, as shown in Table \ref{tbl:case-competition-int-ext-typology} (repeated from Section \ref{sec:pattern-i}). The variation between the four patterns arises from allowing either the internal case, the external case, both of them or neither of them to surface.

\begin{table}[H]
  \center
  \caption{Relative pronoun follows case competition}
  \begin{tabular}{c|c|c|c}
    \toprule
    \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
           & [\ac{nom}]
           & [\ac{acc}]
           & [\ac{dat}]
           \\ \cmidrule{1-4}
       [\ac{nom}]
           & \ac{nom}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{acc}]
           & \ac{acc}
           & \ac{acc}
           & \ac{dat}
           \\ \cmidrule{1-4}
       [\ac{dat}]
           & \ac{dat}
           & \ac{dat}
           & \ac{dat}
           \\
     \bottomrule
  \end{tabular}
    \label{tbl:case-competition-int-ext-typology}
\end{table}

In this section I introduced two different ways of filling out the table. The first one is the one in which the relative pronoun appears in the internal case, as in Table \ref{tbl:no-case-competition-int-typology} (repeated from Table \ref{tbl:no-case-competition-ext}).

\begin{table}[H]
  \center
  \caption{Relative pronoun in internal case}
  \begin{tabular}{c|c|c|c}
    \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{nom}
          & \ac{nom}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{acc}
          & \ac{acc}
          & \ac{acc}
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \ac{dat}
          & \ac{dat}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
  \label{tbl:no-case-competition-int-typology}
\end{table}

The second one is the one in which the relative pronoun appears in the external case, as in Table \ref{tbl:no-case-competition-ext-typology} (repeated from Table \ref{tbl:no-case-competition-ext}).

\begin{table}[H]
  \center
  \caption{Relative pronoun in external case}
  \begin{tabular}{c|c|c|c}
    \toprule
   \textsubscript{\ac{int}} \textsuperscript{\ac{ext}}
          & [\ac{nom}]
          & [\ac{acc}]
          & [\ac{dat}]
          \\ \cmidrule{1-4}
      [\ac{nom}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{acc}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\ \cmidrule{1-4}
      [\ac{dat}]
          & \ac{nom}
          & \ac{acc}
          & \ac{dat}
          \\
    \bottomrule
  \end{tabular}
  \label{tbl:no-case-competition-ext-typology}
\end{table}

I incorporate the parameters that generates these different patterns into the diagram from Section \ref{sec:summary-3-patterns} in Figure \ref{fig:typology}.
I added two different parameters. First, a language either has case competition or it does not at at `case competition?'. If the language has case competition, the pattern shown in Table \ref{tbl:case-competition-int-ext-typology} is generated. The two parameters that follow then (`\ac{int} as winner?' and `\ac{ext} as winner?') are described in Section \ref{sec:summary-3-patterns}.
If the language does not have case competition, the second parameter is whether the language lets its relative pronouns appear either in the internal case or in the external case at at `\ac{int}/\ac{ext}?'.
If the language lets its relative pronouns appear in the internal case, the pattern shown in Table \ref{tbl:no-case-competition-int-typology} is generated. I am not aware of any language that lets its relative pronoun appear in the internal case.\footnote{
In this dissertation I do not offer an explanation for why this type of example should be absent. Future research should determine whether this pattern is actually attested, or whether this option should be excluded and how.
}
If the language lets its relative pronouns appear in the external case, the pattern shown in Table \ref{tbl:no-case-competition-ext-typology} is generated. Old English and Modern Greek are two examples of languages that let their relative pronouns appear in the external case.

\begin{figure}[H]
  \centering
    \begin{tikzpicture}[node distance=1.5cm]
      \footnotesize{
    \node (question1) [question]
    {case\\ competition?};
        \node (question4) [question, below of=question1, xshift=-2.5cm, yshift=-1cm]
        {\ac{int}/\ac{ext}?};
            \node (outcome5) [outcome, below of=question4, xshift=-1.5cm]
            {\ac{int}};
                \node (example5) [example, below of=outcome5, yshift=0.25cm]
                {\scriptsize{n.a. (1)\\\phantom{x}}};
            \node (outcome1) [outcome, below of=question4, xshift=1.5cm]
            {\ac{ext}};
                \node (example1) [example, below of=outcome1, yshift=0.25cm]
                {\scriptsize{e.g. Old English, Modern Greek (5)}};
        \node (question2) [question, below of=question1, xshift=2.5cm, yshift=-1cm]
        {\ac{int} as winner?};
            \node (outcome2) [outcome, below of=question2, xshift=-1.5cm]
            {*};
                \node (example2) [example, below of=outcome2, yshift=0.25cm]
                {\scriptsize{e.g. Polish (9)\\\phantom{x}}};
            \node (question3) [question, below of=question2, xshift=2cm, yshift=-0.5cm]
            {\ac{ext} as winner?};
                \node (outcome3) [outcome, below of=question3, xshift=-1.5cm]
                {\ac{int} + complex};
                    \node (example3) [example, below of=outcome3, yshift=0.25cm]
                    {\scriptsize{e.g. Modern German (3)\\\phantom{x}}};
                \node (outcome4) [outcome, below of=question3, xshift=1.5cm]
                {complex};
                    \node (example3) [example, below of=outcome4, yshift=0.25cm]
                    {\scriptsize{e.g. Gothic, Old High German, Classical Greek (2)}};

    \draw [arrow] (question1) -- node[anchor=west] {yes} (question2);
    \draw [arrow] (question2) -- node[anchor=east] {no} (outcome2);
    \draw [arrow] (question2) -- node[anchor=west] {yes} (question3);
    \draw [arrow] (question3) -- node[anchor=east] {no} (outcome3);
    \draw [arrow] (question3) -- node[anchor=west] {yes} (outcome4);
    \draw [arrow] (question1) -- node[anchor=east] {no} (question4);
    \draw [arrow] (question4) -- node[anchor=east] {\tsc{int}} (outcome5);
    \draw [arrow] (question4) -- node[anchor=west] {\tsc{ext}} (outcome1);
    }
    \end{tikzpicture}

    \caption{Attested patterns in headless relatives}
    \label{fig:typology}
\end{figure}

In Table \ref{tbl:possible-headless-relatives}, I give all logically possible patterns for headless relatives. The top row sketches two different situations: one in which the internal case is the most complex ([\ac{int}]>[\ac{ext}]) and one in which the external case is the most complex ([\ac{ext}]>[\ac{int}]). The second row refers to the case which the relative pronoun appears in, which can be either the internal case (\ac{int}) or the external case (\ac{ext}).

When the internal case and the external case differ (which holds for both options the top row indicates), the relative pronoun cannot appear in both the internal and external case at the same time. This excluded the possibility of having a checkmark at both \ac{int} and \ac{ext} in the same situation. This leaves the possibility to have a checkmark at \ac{int}, at \ac{ext} or at none of them. This gives 3 × 3 = 9 logically possible options, which are listen in Table \ref{tbl:possible-headless-relatives}.

\begin{table}[H]
  \center
  \caption{Possible patterns for headless relatives}
    \begin{tabular}{ccc|ccc}
    \toprule
      &   \multicolumn{2}{c}{[\ac{int}]>[\ac{ext}]} & \multicolumn{2}{|c}{[\ac{ext}]>[\ac{int}]} &                  \\
          \cmidrule(lr){2-3}                      \cmidrule(lr){4-5}
      &   \ac{int}            & \ac{ext}          & \ac{int}          & \ac{ext}            & language              \\
          \cmidrule(lr){2-2}  \cmidrule(lr){3-3}  \cmidrule(lr){4-4}  \cmidrule(lr){5-5}    \cmidrule(lr){6-6}
    1 &   ✔                   & *                 & ✔               & *                     & n.a.                  \\
    2 &   ✔                   & *                 & *               & ✔                     & e.g. Old High German  \\
    3 &   ✔                   & *                 & *               & *                     & e.g. Modern German    \\
    4 &   {*}                 & ✔                 & ✔               & *                     & n.a.                  \\
    5 &   {*}                 & ✔                 & *               & ✔                     & e.g. Old English      \\
    6 &   {*}                 & ✔                 & *               & *                     & n.a.                  \\
    7 &   {*}                 & *                 & ✔               & *                     & n.a.                  \\
    8 &   {*}                 & *                 & *               & ✔                     & n.a.                  \\
    9 &   {*}                 & *                 & *               & *                     & e.g. Polish           \\
    \bottomrule
  \end{tabular}
    \label{tbl:possible-headless-relatives}
\end{table}

In what follows I show how Figure \ref{fig:typology} generates of all logically possible patterns only the attested patterns (except for the one in which the relative pronoun always takes the internal case).

I start with the leftmost pattern in Figure \ref{fig:typology}, which is number 1 in Table \ref{tbl:possible-headless-relatives}. In this pattern, there is no case competition, and the relative pronoun surfaces in the internal case. As I mentioned earlier, I am not aware of a language that exemplified this pattern and future research should tell whether this option is attested or whether it should be excluded.
The second pattern in Figure \ref{fig:typology} is number 5 in Table \ref{tbl:possible-headless-relatives}. In this pattern, there is no case competition, and the relative pronoun surfaces in the external case. This pattern is exemplified by Old English and Modern Greek.
The third pattern in Figure \ref{fig:typology} is number 9 in Table \ref{tbl:possible-headless-relatives}. In this pattern, there is case competition, and the relative pronoun is neither allowed to surface in the internal case nor in the external case when either of them wins the case competition. This pattern is exemplified by Polish.
The fourth pattern in Figure \ref{fig:typology} is number 3 in Table \ref{tbl:possible-headless-relatives}. In this pattern, there is case competition, and the relative pronoun is only allowed to surface in the internal case when it wins the case competition. This pattern is exemplified by Modern German.
The fifth and last pattern in Figure \ref{fig:typology} is number 2 in Table \ref{tbl:possible-headless-relatives}. In this pattern, there is case competition, and the relative pronoun is allowed to surface in the internal case and the external case when either of them wins the case competition. This pattern is exemplified by Old High German, Gothic and Classical Greek.

This leaves four patterns that are logically possible but not attested in languages: pattern numbers 4, 6, 7 and 8 in Table \ref{tbl:possible-headless-relatives}. These patterns cannot be generated by the diagram in Figure \ref{fig:typology}. That means that they are not a result of any of the possible parameter settings in the diagram.

In the pattern number 4, the relative pronoun surfaces in the external case when the internal case is the most complex, and the relative pronoun surfaces in the internal case when the external case is the most complex. In other words, the relative pronoun  appears in the losing case in the case competition.
Pattern number 6 and 7 are both subsets of pattern number 4 in the sense that they allow part of what number 4 allows.
In the pattern number 6, the relative pronoun surfaces in the external case when the internal case is the most complex, and there is no grammatical option when the external case is the most complex.
Patterns number 7 is the opposite of pattern number 6: there is no grammatical option when the external case is the most complex, and the relative pronoun surfaces in the internal case when the external case is the most complex.
The absence of these three patterns across languages provides further evidence for the case scale in Chapter \ref{ch:recurring}.

In the pattern number 8, the relative pronoun is only allowed to surface in the external case when it wins the case competition. This pattern is excluded as a result of the relative ordering of `\ac{int} as a winner?' and `\ac{ext} as a winner?' in the diagram in Figure \ref{fig:typology}. The next chapter, Chapter \ref{ch:decomposition}, discusses the linguistic counterpart of this ordering.
