% !TEX root = thesis.tex

\chapter{\tsc{d} vs. \tsc{wh}}\label{ch:relativization}

Proposal of wh in a nutshell
Proposal of d in a nutshell

% Some intro about head of the relative clause and relative pronoun.


% So, we have seen that relative pronouns are sensitive to the internal case and to the external case. In this chapter we go back to what this means for the `head' in headless relatives. In headless relatives this head is namely missing. Like I said in the introduction, there are several ways of interpreting that.

% In this section I discuss the matter whether it is actually the relative pronoun or the head of the relative clause that surfaces in headless relatives. I argue that for Modern German it is always the relative pronoun in the relative clause that surfaces. This follows from the fact that Modern German has \tsc{wh}-pronouns as relative pronouns in its headless relatives. In Gothic and Old High German it is sometimes the relative pronoun and sometimes the (light) head of the relative clause that surfaces as relative pronoun. This is a consequence of the \tsc{d}-morphology of the relative pronoun in Old High German and Gothic.

% I connect this behavior of Modern German headless relatives to the shape of its relative pronouns. Modern German has \tsc{wh}-pronouns as relative pronouns in headless relative constructions. In this section I show that it is a consequence of having \tsc{wh}-pronouns as relative pronouns that the internal case can only surface as winner in the case competition. The different behavior of Gothic and Old High German headless relatives follows from the \tsc{d}-pronoun these languages use as their relative pronouns.

% In both types of languages, I analyze it as if there was a head, but it is phonologically empty/deleted. I show that this allows me to make a distinction between German and languages like Polish, that do not allow for the case competition as German does. OHG and Gothic also independently show that they can be derived from a double element structure.




\section{\tsc{wh}-relatives}

My claim: the behavior in Moden German headless relatives is a consequence of its \tsc{wh}-morphology.

In this section I discuss headless relatives in Modern German. Modern German only allows the relative pronoun only to surface in the internal case, not in the external case. The relative pronoun used in headless relatives in this language is a \tsc{wh}-pronoun. I show in this section that syntactically it is part of the relative clause but that the relative pronoun is sensitive to both the internal and external case.

Modern German differs crucially from Gothic and Old High German in that its headless relative are only grammatical when the internal case wins the case competition over the external case. In other words, the relative pronoun can only surface in the internal case, i.e. the case of the relative clause. This seems to indicate that the relative pronoun is part of the relative clause.

This can be modelled as follows:

 \ex.
 \begin{forest} boom
		 	[CP,no edge, s sep=20mm
	       [\ac{acc}P,
				 tikz={
				 \node[label=below:\tit{wen},
				 draw,circle,
				 scale=0.85,
				 fit to=tree]{};
				 }
	           [\tsc{f2}]
	           [\tsc{nomP}
	               [\tsc{f1}]
	               [XP
	                   [\phantom{xxx}, roof]
	               ]
	           ]
	       ]
				 [VP
				 		 [\tit{Maria mag}, roof]
				 ]
			]
 \end{forest}

The structure shows an accusative in the relative clause. An \tsc{acc}P by definition contains a \tsc{nom}P, so the nominative is there.

We also understand that the other way around does not work: a nominative in the relative clause does not contain an accusative.


\subsection{The relative pronoun in the relative clause}

Independent from these case facts, I provide three arguments why the headless relative pronoun in Modern German is part of the relative clause. The first one is extraposition in headed relative clauses, the second one is from headless relatives, the last one is \tsc{wh}-elements in general.

The strongest argument for the headless relative pronoun being part of the relative clause comes from extraposition in headless relatives. In Modern German, it is possible to extrapose a clause (a CP), but not a noun phrase (DP). The relative clause including the relative pronoun in a headless relative can be extraposed, which indicates that the relative pronoun is part of the relative clause. In what follows I show that CPs can be extraposed and DPs cannot. Then I illustrate how relative clauses including the relative pronoun in headless relatives pattern with CPs.

The sentence in \ref{ex:mg-extrapose-cp} show that it is possible to extrapose CP. In \ref{ex:mg-extrapose-cp-base}, the clausal object \tit{wie es dir geht} `how you are doing', marked here in bold, appears in its base position. This objects can be extraposed to the right edge of the clause, shown in \ref{ex:mg-extrapose-cp-moved}.

\ex.\label{ex:mg-extrapose-cp}
\ag. Mir ist \tbf{wie} \tbf{es} \tbf{dir} \tbf{geht} egal.\\
 1\tsc{sg}.\tsc{dat} is how it 2\tsc{sg}.\tsc{dat} goes {the same}\\
 `I don't care how you are doing.' \flushfill{Modern German}\label{ex:mg-extrapose-cp-base}
\bg. Mir is egal \tbf{wie} \tbf{es} \tbf{dir} \tbf{geht}.\\
 1\tsc{sg}.\tsc{dat} is {the same} how it 2\tsc{sg}.\tsc{dat} goes\\
 `I don't care how you are doing.' \flushfill{Modern German}\label{ex:mg-extrapose-cp-moved}

\ref{ex:mg-extrapose-dp} illustrates that it is impossible to extrapose a DP. The clausal object of \ref{ex:mg-extrapose-cp} is replaced by a simplex noun phrase \tit{die Sache} `that matter'.
In \ref{ex:mg-extrapose-dp-base} the object, marked in bold, appears in its base position. In \ref{ex:mg-extrapose-dp-moved} it is extraposed, and the sentence is no longer grammatical.

\ex.\label{ex:mg-extrapose-dp}
\ag. Mir ist \tbf{die} \tbf{Sache} egal.\\
 1\tsc{sg}.\tsc{dat} is that matter {the same}\\
 `I don't care about that matter.' \flushfill{Modern German}\label{ex:mg-extrapose-dp-base}
\bg. *Mir ist egal \tbf{die} \tbf{Sache}.\\
 1\tsc{sg}.\tsc{dat} is {the same} that matter\\
 `I don't care about that matter.' \flushfill{Modern German}\label{ex:mg-extrapose-dp-moved}

So here we see that with an over head.


 \ex. \label{ex:extra-headed} adapted from Groos/v Riemsdijk
 \ag. Jan hat das, \tbf{was} \tbf{er} \tbf{gestohlen} \tbf{hat}, zurückgegeben.\\
 Jan has the money which he stolen has returned\\
 \glt `Jan has returned the money that he has stolen.'\label{ex:extra-headed-base}
 \bg. Jan hat das zurückgegeben, \tbf{was} \tbf{er} \tbf{gestohlen} \tbf{hat}.\\
 the Hans has the money returned which he stolen has\\
 \glt `Jan has returned the money that he has stolen.'\label{ex:extra-headed-only-clause}
 \cg. *Jan hat zurückgegeben, das, \tbf{was} \tbf{er} \tbf{gestohlen} \tbf{hat}.\\
 Jan has returned the money which he stolen has\\
 \glt `Jan has returned the money that he has stolen.'\label{ex:extra-headed-head-clause}

Here without the head.
\ref{ex:mg-extra} shows an example of a headless relative. The relative clause including the relative pronoun, marked in bold, appears in its base position in \ref{ex:mg-extra-no}. In \ref{ex:mg-extra-yes}, it is extraposed.

\ex.\label{ex:mg-extra}
\ag. Jan hat \tbf{was} \tbf{er} \tbf{gestohlen} \tbf{hat} zurückgegeben.\\
the Hans has what he stolen has returned\\
`Hans has returned what he has stolen.' \citet[185]{groos1981}\label{ex:mg-extra-no}
\bg. Jan hat zurückgegeben \tbf{was} \tbf{er} \tbf{gestohlen} \tbf{hat}.\\
the Hans has returned what he stolen has\\
`Hans has returned what he has stolen.' \citet[185]{groos1981}\label{ex:mg-extra-yes}

\ref{ex:mg-extrapose-cp} and \ref{ex:mg-extrapose-dp} have shown that only CPs can be moved in Modern German and DPs cannot. \ref{ex:mg-extra} shows that the relative clause including the relative pronoun can be extraposed, so it is a CP.

The second argument for the relative pronoun in headless relative being part of the relative clause comes from relative pronouns in headed relatives. Even though these here you can see too that the head is a DP in the main clause and the relative pronoun is part of the relative clause.

\ref{ex:extra-headed-base} gives an example of a headed relative clause with the relative clause in base position. \tit{das Geld} `the money' is the head of the relative clause an \tit{das er gestohlen hat} `that he stole' is the relative clause.
\ref{ex:extra-headed-only-clause} gives an example with the relative clause extraposed, so at the right edge of the clause. The head of the relative clause is left behind. This is grammatical. This means that a relative clause can be extraposed.
\ref{ex:extra-headed-head-clause} gives an example with the relative caluse extraposed and the head of the relative clause with it. This is ungrammatical in Modern German. This means that the head of the relative clause cannot be extraposed.


%
% Last, I consider \tsc{wh}-pronouns in Modern German in general. There are more \tsc{wh}-pronouns, such as \tsc{wh}-interrogatives (but not indefinites), that typically appear at the left edge of the clause, just like in other \tsc{wh}-moving languages. An example with \tit{wen} `who' is given in \ref{ex:german-wh-left}. Non-\tsc{wh}-pronouns do not necessarily appear at the left edge of the clause, just as \tit{ihn} `him' in \ref{ex:german-non-wh-not-left}.
%
% \ex.
% \ag. Wen hast du gestern gesehen?\\
%  Who have you yesterday seen\\
%  `Who did you see yesterday?'\label{ex:german-wh-left}
% \bg. Ich habe ihn gestern gesehen.\\
%  I have him yesterday seen\\
%  `I saw him yesterday.'\label{ex:german-non-wh-not-left}

Conclusion: the \tsc{wh}-pronoun is part of the relative clause in Modern German. It appears at the left edge.



\subsection{Two problems}

(1) how does this licensing take place? Normally we have a single DP satisfying the case of a single V. What is happening here?
(2) if we follow this logic, to show this type of case competition? (OHG and Gothic also allow this one, so that is fine, they just allow something else in addition, and I'll get back to that later) But there are languages like Polish, which is so-called `strictly matching'. It always gives an ungrammatical result when there is case competition (and not syncretism or two identical cases).


\subsubsection{Care about external case}

%%relative pronoun cares about internal and external case
However, just like in Gothic and Old High German, the Modern German relative pronoun is sensitive to the internal and external case.
The internal case only surfaces when it is more complex than the external case.

Consider the examples in \ref{ex:mg-internal}. In both sentences, the internal case is accusative, because the predicate in the relative clause \tit{mögen} `to like' takes accusative objects. The external case differs between the two sentences. In \ref{ex:mg-dat-acc-wen} the external case is dative, because the predicate \tit{vertrauen} `to trust' takes dative objects.  In \ref{ex:mg-nom-acc-wen}, the external case is nominative, because \tit{besuchen} `to visit' takes nominative subjects.

\ex.\label{ex:mg-internal}
\ag. *Ich vertraue wen \tbf{auch} \tbf{Maria} \tbf{mag}. \\
I.\ac{nom} trust.1\ac{sg}\scsub{[dat]} \tsc{rel}.\ac{acc}.\tsc{an} also Maria.\ac{nom} like.3\ac{sg}\scsub{[acc]}.\\
`I trust whoever Maria also likes.' \flushfill{adapted from \pgcitealt{vogel2001}{345}}\label{ex:mg-dat-acc-wen}
\bg. Uns besucht \tbf{wen} \tbf{Maria} \tbf{mag}.\\
 we.\ac{acc} visit.3\ac{sg}\scsub{[nom]} \tsc{rel}.\ac{acc}.\tsc{an} Maria.\ac{nom} like.3\ac{sg}\scsub{[acc]}\\
 `Who visits us, Maria likes.' \flushfill{adapted from \pgcitealt{vogel2001}{343}}\label{ex:mg-nom-acc-wen}

The sentence in \ref{ex:mg-dat-acc-wen} is ungrammatical, and the one in \ref{ex:mg-nom-acc-wen} is not. The relative clauses in both sentences are identical case-wise, so the difference in ungrammaticality cannot lie there. The only observable differences are found in the main clause. The former sentence is ungrammatical, because the internal accusative case cannot win the case competition over the external dative. The latter sentence is grammatical, because the internal accusative case wins over the external nominative. The conclusion that can be drawn from this is that the relative pronoun in Modern German headless relatives cares about both the internal and the external case.

In sum, even though the relative pronoun in Modern German headless relatives is always part of the relative clause, the relative pronoun also takes the external case into account. That means that the relative pronoun somehow needs to have access to the main clause case. I return to the matter of `somehow' later on in this section.





\subsubsection{Other \tsc{wh}-languages}

\ex.
\ag. Jan lubi \tbf{komu} \tbf{(kolwiek)} \tbf{dokucza}.\\
Jan like.3\tsc{sg}\scsub{[acc]} \tsc{rel}.\tsc{dat}.\tsc{sg}.\tsc{an} ever tease.3\tsc{sg}\scsub{[dat]}\\
`Jan likes whoever he teases.'
\bg. Jan ufa \tbf{komu} \tbf{(kolwiek)} \tbf{wpuścil} \tbf{do} \tbf{domu}.\\
 Jan trust.3\tsc{sg}\scsub{[dat]} \tsc{rel}.\tsc{dat}.\tsc{sg}.\tsc{an} ever let.3\tsc{sg}\scsub{[acc]} to home\\
 `Jan trusts whoever he let into the house.'





\subsection{Solution: an external head}

I propose a single answer to both of these questions: there is an external head. This is the head of the relative clause, and it is deleted in Modern German (as long as it is smaller than the relative pronoun). This external head in Polish is not deleted: it does not add any extra definiteness or indefiniteness. Before I show how this works exactly, I show some independent evidence for assuming that there is this head.


\subsubsection{Languages with two heads}

There is also independent evidence for this head, namely from languages that actually let the head surface.

\exg. doü adiyano-no \tbf{doü} deyalukhe\\
 sago give.3\tsc{pl}.\tsc{nonfut}-\tsc{conn} sago finished.\tsc{ajd}\\
 `The sago that they gave is finished.' \flushfill{Kombai, Dryer 2005}

\exg. gana gu fali-kha \tbf{ro} na-gana-y-a\\
 {bush knife} 2\tsc{sg} carry-go.\tsc{2sg}.\tsc{nonfut} my-{bush knife}-\tsc{tr-pred}\\
 `The bush knife that you took away, is my bush knife.'

\exg. yare gamo khereja bogi-n-o \tbf{rumu} na-momof-a\\
 {old man} join.\tsc{ss} work \tsc{dur}.do.\tsc{3sg}.\tsc{nf}-\tsc{tr}-\tsc{conn} person my-uncle-\tsc{pred}\\
 `The old man who is joining the work is my uncle.'

\ex.
\ag. Junya-wa[Ayaka-gamui-taringo]-otabe-ta.\\
 Junya-topAyaka-nompeel-pastapple-acceat-past\\
 ‘Junya ate the apples that Ayaka peeled.’
\bg. Junya-wa[Ayaka-garingo-omui-ta-no]-otabe-ta.\\
 Junya-topAyaka-nomapple-accpeel-past-no-acceat-past\\
 literally ‘Junya ate [that Ayaka peeled apples].’
\bg. Junya-wa[Ayaka-garingo-omui-tasono-ringo]-otabe-ta.\\
 Junya-topAyaka-nomapple-accpeel-pastthat-apple-acceat-past\\
 literally ‘Junya ate [those apples that Ayaka peeled apples].’

So, we have the head.


\subsubsection{Solving external case care}

--give here a table with on one side the tree repeated and on the other side the head, with XP at the bottom?-- %split the relative pronoun up in the w part and the other part?

Now this accusative here can license an external nominative case. The idea is: it is inside this accusative already. We understand why it could not license a dative: \tsc{f3} is not contained in the accusative.


\subsubsection{Solving other \tsc{wh}-languages}

But if we follow the logic I just sketched, I expect all languages that have \tsc{wh}-pronouns as relative pronouns to allow for letting the internal case win in case competition. This prediction is incorrect.

Under the grafting analysis, there is nothing that distinguishes Polish from Modern German, so there is no reason for the Polish example in X to be incorrect. The other one is ungrammatical, just as the German one, but the first one is too, and that is unexpected.

Now I do not want to say that Polish and German differ in that Polish is `strictly matching' and Modern German is `more relaxed', that this is something construction- and language-specific. Instead, I let the distinction in headless relatives follow from something within the language. This something is their light-headed relatives.

\ex. Polish light-headed relative

\ex. German light-headed relative
\a. das was
\b. das das

In German a sense of definiteness is added, because of the \tsc{d} in \tit{das}. In Polish that is not the case, because \tit{to} in Polish does not necessarily have definiteness. Evidence for that for Czech comes from Radek. I take that to exist for Polish as well.

So, Polish can involve a second head (the light head) without changing the meaning of the construction. German cannot. Now it is important to note the timing of this `repair' strategy. This has to namely be in the course of the derivation, it cannot be something that is inserted `afterwards'. What I mean with this is that there needs to be an element (which is going to become the light head in Polish but not in German) that is available during the derivation. Depending on what this element looks like, Polish shows the light head.

\ex. Polish
\a. rel clause with \tsc{acc}
\b. other element: \tsc{nom}
\b. t, c

\ex. German
\a. rel clause with \tsc{acc}
\b. other element: \tsc{nom}
\b. w (because d would change the meaning)

So, Polish has an `out', which German does not have.

What is this external element? This is the external head, that actually shows up in some languages: double-headed relative clauses.



\subsection{Syntactic position of external head}

Where is this head in the syntactic structure?

  \begin{itemize}
    \item Somewhere where the relative pronoun can delete it: where it is c-commanded by the relative pronoun
    \item Somewhere where it can receive case from the main clause
    \item Where it normally is in SOV languages (does the thing in Polish move because it is a svo language?)
  \end{itemize}


So this works.

\ex.
\begin{forest} boom
[, s sep=20mm
    [CP, s sep=20mm
        [\ac{acc}P,
        tikz={
        \node[label=below:\tit{wen},
        draw,circle,
        scale=0.85,
        fit to=tree]{};
        }
            [\tsc{f2}]
            [\tsc{nomP},
            tikz={
            \node[draw,circle,transparent,
            fill=DG,fill opacity=0.2,
            scale=0.8,
            fit to=tree]{};
            }
                [\tsc{f1}]
                [XP
                    [\phantom{xxx}, roof]
                ]
            ]
        ]
        [VP
            [\tit{Maria mag}, roof]
        ]
    ]
    [\textcolor{LG}{\tsc{nomP}},
    tikz={
    \node[draw,circle,
    scale=0.8,
    fit to=tree]{};
    }
        [\textcolor{LG}{\ac{nom}},edge=LG]
        [\textcolor{LG}{XP},baseline,edge=LG
            [\textcolor{LG}{\phantom{xxx}},
            roof, baseline, edge=LG
            ]
        ]
    ]
]
\end{forest}

But here it does not.

\ex.
\begin{forest} boom
[, s sep=20mm
    [CP, s sep=20mm
        [\ac{nom}P,
        tikz={
        \node[label=below:\tit{wer},
        draw,circle,
        fill=DG,fill opacity=0.2,
        scale=0.85,
        fit to=tree]{};
        }
            [\tsc{f1}]
            [XP
                [\phantom{xxx}, roof]
            ]
        ]
        [VP
            [\tit{mir sympatisch ist}, roof]
        ]
    ]
    [\ac{acc}P
        [\tsc{f2}]
        [\textcolor{LG}{\tsc{nomP}},
        tikz={
        \node[draw,circle,
        scale=0.8,
        fit to=tree]{};
        }
            [\textcolor{LG}{\ac{nom}},edge=LG]
            [\textcolor{LG}{XP},baseline,edge=LG
                [\textcolor{LG}{\phantom{xxx}},
                roof, baseline, edge=LG
                ]
            ]
        ]
    ]
]
\end{forest}



\tit{as} is deleted because it cannot surface on its own? why not?

what is this deletion to begin with?

Finnish is like German because its light-headed relatives also add some definiteness.




\section{\tsc{d}-relatives}

My claim: d-morphology allows for both directions

\subsection{Like Modern German}

d-relative pronoun is also in embedded clause
same story

\subsection{Derived from light-headed relatives}

main clause more complex: full double-headed structure, deletion of element in speccp

Why does this not happen in Modern German? no idea..

\subsubsection{Old High German}

has attraction, so it could be derived from deletion under c-command under identity

\subsubsection{Gothic}

does not have attraction








\section{Alternative analyses}

\subsection{Himmelreich}



\subsection{Grafting story}

For this pattern a single element analysis seems intuitive, if you assume that case is complex and that syntax works bottom-up. First you built the relative clause, with the big case in there. Then you build the main clause and you let the more complex case in the embedded clause license the main clause predicate.

Consider the example in \ref{ex:mg-nom-acc-grafting}. Here the internal case is accusative and the external one nominative.

\exg. Uns besucht \tbf{wen} \tbf{Maria} \tbf{mag}.\\
 we.\ac{acc} visit.3\ac{sg}\scsub{[nom]} \tsc{rel}.\ac{acc}.\tsc{an} Maria.\ac{nom} like.3\ac{sg}\scsub{[acc]}\\
 `Who visits us, Maria likes.' \flushfill{adapted from \pgcitealt{vogel2001}{343}}\label{ex:mg-nom-acc-grafting}

The relative clause is built, including the accusative relative pronoun. Now the main clause predicate can merge with the nominative that is contained within the accusative.

 \ex.
 \begin{forest} boom
	 [,name=src, s sep=15mm
			[VP
			 		[\tit{besucht}, roof]
			]
		 	[,no edge, s sep=20mm
	       [\ac{acc}P,
				 tikz={
				 \node[label=below:\tit{wen},
				 draw,circle,
				 scale=0.85,
				 fit to=tree]{};
				 }
	           [\tsc{f2}]
	           [\tsc{nomP},name=tgt
	               [\tsc{f1}]
	               [XP
	                   [\phantom{xxx}, roof]
	               ]
	           ]
	       ]
				 [VP
				 		 [\tit{Maria mag}, roof]
				 ]
			]
	 ]
	 \draw (src) to[out=south east,in=north east] (tgt);
 \end{forest}\label{ex:acc-nom-grafting}

The other way around does not work. Consider \ref{ex:mg-acc-nom-grafting}. This is an example with nominative as internal case and accusative as external case.

\exg. *Ich {lade ein}, wen \tbf{mir} \tbf{sympathisch} \tbf{ist}.\\
I.\ac{nom} invite.1\ac{sg}\scsub{[acc]} \tsc{rel}.\ac{acc}.\tsc{an} I.\ac{dat} nice be.3\ac{sg}\scsub{[nom]}\\
`I invite who I like.' \flushfill{adapted from \pgcitealt{vogel2001}{344}}\label{ex:mg-acc-nom-grafting}

Now the relative clause is built first again, this time only including the nominative case. There is no accusative node to merge with for the external predicate. Instead, the relative pronoun would need to grow to accusative somehow and then the merge could take place. This is the desired result, because the sentence is ungrammatical.

\ex.
\begin{forest} boom
  [,name=src, s sep=15mm
     [VP
         [\tit{lade ein}, roof]
     ]
         [,no edge
    			[\tsc{nomP},
    			tikz={
    			\node[label=below:\tit{wer},
    			draw,circle,
    			scale=0.85,
    			fit to=tree]{};
    			}
    					[\tsc{f1}]
    					[XP
    							[\phantom{xxx}, roof]
    					]
    			]
    			[VP
    					[\tit{mir sympatisch ist}, roof]
    			]
    	 ]
    ]
\end{forest}\label{ex:nom-acc-grafting}

So, this seems to work fine. The assumptions you have to do in order to make this are the following. First, case is complex. Second, you can remerge an embedded node (grafting). For the first one I have argued in Chapter \ref{ch:decomposition}. The second one could use some additional argumentation. It is a mix between internal remerge (move) and external merge, namely external remerge. Other literature on multidominance and grafting, other phenomena. Problems: linearization, .. But even if fix all these theoretical problems, there is an empirical one.

That is, I want to connect this behavior of Modern German headless relatives to the shape of its relative pronouns. These pronouns are \tsc{wh}-elements. The OHG and Gothic ones are not \tsc{wh}, they are \tsc{d}. Their relative pronouns look different, and so their headless relatives can also behave differently.
