% !TEX root = thesis.tex

\chapter{Beyond Germanic}

Greek, Finnish, Polish

\section{Polish (\citealt[17]{himmelreich2017}, after \citealt{citko2013})}

\ex.
\ag. *Jan ufa kogokolkwiek wpu{\'s}cil do domu.\\
Jan trusts\scsub{dat} who.\tsc{acc} let\scsub{acc} to home\\
`Jan trusts whoever he let into the house.'
\bg. *Jan ufa komukolkwiek wpu{\'s}cil do domu.\\
Jan trusts\scsub{dat} who.\tsc{dat} let\scsub{acc} to home\\
`Jan trusts whoever he let into the house.'

\ex.
\ag. *Jan lubi kogokolkwiek dokucza.\\
Jan likes\scsub{acc} who.\tsc{acc} teases\scsub{dat}\\
`Jan likes whoever he teases.'
\bg. *Jan lubi komukolkwiek dokucza.\\
Jan likes\scsub{acc} who.\tsc{dat} teases\scsub{dat}\\
`Jan likes whoever he teases.'

\begin{table}[h]
	\center
	\setlength{\tabcolsep}{10pt}
	\begin{tabular}{|c|c|c|c|c|}\hline
		\diagbox{m}{e}	& V\scsub{nom} 											& V\scsub{acc} 											& V\scsub{gen} 											& V\scsub{dat} 											\\\hline
		V\scsub{nom} 		& \tsc{nom} 												& \diagbox{*\tsc{nom}}{*\tsc{acc}}	& \diagbox{*\tsc{nom}}{*\tsc{gen}}	& \diagbox{*\tsc{nom}}{*\tsc{dat}} 	\\\hline
		V\scsub{acc}		& \diagbox{*\tsc{acc}}{*\tsc{nom}}	& \tsc{acc} 												& \tsc{acc/gen} 										& \diagbox{*\tsc{acc}}{*\tsc{dat}} 	\\\hline
		V\scsub{gen} 		& \diagbox{*\tsc{gen}}{*\tsc{nom}}	& \tsc{acc/gen} 										& \tsc{gen} 												& \diagbox{*\tsc{gen}}{*\tsc{dat}} 	\\\hline
		V\scsub{dat} 		& \diagbox{*\tsc{dat}}{*\tsc{nom}}	&	\diagbox{*\tsc{dat}}{*\tsc{acc}}	& \diagbox{*\tsc{dat}}{*\tsc{gen}} 	& \tsc{dat}													\\\hline
	\end{tabular}
\end{table}



\section{Greek}

\ex. free relative pronoun takes embedded clause case
\ag. *Irϑan ópjus káleses.\\
came.3\tsc{pl}\scsub{nom} who.\tsc{acc.pl} invited.2\tsc{sg}\scsub{acc}\\
`Whoever you invited came.' \hfill m:\tsc{nom}, e:\tsc{acc}
\bg. *Me efχarístisan ópjon íχa ósi leftá.\\
\tsc{cl.1sg.acc} thanked.3\tsc{pl}\scsub{nom} who.\tsc{gen.pl} had.1\tsc{sg}\scsub{gen} given money\\
`Whoever I had given money to thanked me.' \hfill m:\tsc{nom}, e:\tsc{gen}
\bg. *Efχarístisa ópji me voíϑisan.\\
thanked.1\tsc{sg}\scsub{acc} who.\tsc{nom.pl} \tsc{cl.1sg.acc} helped.3\tsc{pl}\scsub{nom}\\
`I thanked whoever helped me.' \hfill m:\tsc{acc}, e:\tsc{nom}
\bg. *Γnórisa ópju éδosan tin ipotrofia.\\
met.1\tsc{sg}\scsub{acc} who.\tsc{gen.sg} gave.3\tsc{pl}\scsub{gen} the scholarship.\tsc{acc}\\
`I met whoever they gave the scholarship to.' \hfill m:\tsc{acc}, e:\tsc{gen}
\bg. *Eδósa leftá ópjos me voíϑise.\\
gave.1\tsc{sg}\scsub{gen} money who.\tsc{nom.sg} \tsc{cl.1sg.acc}  helped.3\tsc{sg}\scsub{nom}\\
`I gave money to whoever helped me.' \hfill m:\tsc{gen}, e:\tsc{nom}
\bg. *Eδósa leftá ópjon γnórisa.\\
gave.1\tsc{sg}\scsub{gen} money who.\tsc{acc.sg} met.1\tsc{sg}\scsub{acc}\\
`I gave money to whoever I met.' \hfill m:\tsc{gen}, e:\tsc{acc}

\begin{table}[h]
	\center
	\begin{tabular}{|c|c|c|c|}\hline
		\diagbox{m}{e}	& \tsc{nom} 	& \tsc{acc} 	& \tsc{gen} 	\\\hline
		\tsc{nom} 			& \tsc{nom} 	& *\tsc{acc} 	& *\tsc{gen} 	\\\hline
		\tsc{acc} 			& *\tsc{nom}	& \tsc{acc} 	& *\tsc{gen} 	\\\hline
		\tsc{gen} 			& *\tsc{nom} 	& *\tsc{acc} 	& \tsc{gen}		\\\hline
	\end{tabular}
\end{table}


\ex. free relative pronoun takes main clause case
\ag. Irϑan ópji káleses.\\
came.3\tsc{pl}\scsub{nom} who.\tsc{nom.pl} invited.2\tsc{sg}\scsub{acc}\\
`Whoever you invited came.' \hfill m:\tsc{nom}, e:\tsc{acc}, PL
\b. Lefteris?
\gll Irϑan ópjos káleses.\\
came.3\tsc{pl}\scsub{nom} who.\tsc{nom.sg} invited.2\tsc{sg}\scsub{acc}\\
`Whoever you invited came.' \hfill m:\tsc{nom}, e:\tsc{acc}, SG
\bg. *Me efχarístisan ópji íχa δósi leftá.\\
\tsc{cl.1sg.acc} thanked.3\tsc{pl}\scsub{nom} who.\tsc{nom.pl} had.1\tsc{sg}\scsub{gen} given money\\
`Whoever I had given money to thanked me.' \hfill m:\tsc{nom}, e:\tsc{gen}, PL
\b. Lefteris?
\gll Me efχarístisan ópjos íχa δósi leftá.\\
\tsc{cl.1sg.acc} thanked.3\tsc{pl}\scsub{nom} who.\tsc{nom.sg} had.1\tsc{sg}\scsub{gen} given money\\
`Whoever I had given money to thanked me.' \hfill m:\tsc{nom}, e:\tsc{gen}, SG
\bg. Efχarístisa ópjus me voíϑisan.\\
thanked.1\tsc{sg}\scsub{acc} who.\tsc{acc.pl} \tsc{cl.1sg.acc} helped.3\tsc{pl}\scsub{nom}\\
`I thanked whoever helped me.' \hfill m:\tsc{acc}, e:\tsc{nom}, PL
\b. Lefteris?
\gll Efχarístisa ópjon me voíϑisan.\\
thanked.1\tsc{sg}\scsub{acc} who.\tsc{acc.sg} \tsc{cl.1sg.acc} helped.3\tsc{pl}\scsub{nom}\\
`I thanked whoever helped me.' \hfill m:\tsc{acc}, e:\tsc{nom}, SG
\bg. *Γnórisa ópjon éδosan tin ipotrofia.\\
met.1\tsc{sg}\scsub{acc} who.\tsc{acc.sg} gave.3\tsc{pl}\scsub{gen} the scholarship.\tsc{acc}\\
`I met whoever they gave the scholarship to.' \hfill m:\tsc{acc}, e:\tsc{gen}, SG
\b. Lefteris?
\gll Γnórisa ópjus éδosan tin ipotrofia.\\
met.1\tsc{sg}\scsub{acc} who.\tsc{acc.pl} gave.3\tsc{pl}\scsub{gen} the scholarship.\tsc{acc}\\
`I met whoever they gave the scholarship to.' \hfill m:\tsc{acc}, e:\tsc{gen}, PL
\bg. Eδósa leftá ópju me voíϑise.\\
gave.1\tsc{sg}\scsub{gen} money who.\tsc{gen.sg} \tsc{cl.1sg.acc}  helped.3\tsc{sg}\scsub{nom}\\
`I gave money to whoever helped me.' \hfill m:\tsc{gen}, e:\tsc{nom}, SG
\b. Lefteris?
\gll Eδósa leftá ópjon me voíϑise.\\
gave.1\tsc{sg}\scsub{gen} money who.\tsc{gen.pl} \tsc{cl.1sg.acc}  helped.3\tsc{sg}\scsub{nom}\\
`I gave money to whoever helped me.' \hfill m:\tsc{gen}, e:\tsc{nom}, PL
\bg. *Eδósa leftá ópju γnórisa.\\
gave.1\tsc{sg}\scsub{gen} money who.\tsc{gen.sg} met.1\tsc{sg}\scsub{acc}\\
`I gave money to whoever I met.' \hfill m:\tsc{gen}, e:\tsc{acc}, SG
\b. Lefteris?
\gll Eδósa leftá ópjon γnórisa.\\
gave.1\tsc{sg}\scsub{gen} money who.\tsc{gen.pl} met.1\tsc{sg}\scsub{acc}\\
`I gave money to whoever I met.' \hfill m:\tsc{gen}, e:\tsc{acc}, PL



\begin{table}[h]
	\center
	\setlength{\tabcolsep}{10pt}
	\begin{tabular}{|c|c|c|c|}\hline
		\diagbox{m}{e}	& V\scsub{nom} 										& V\scsub{acc} 											& V\scsub{gen} 											\\\hline
		V\scsub{nom} 		& \tsc{nom} 											& \diagbox{\tsc{nom}}{*\tsc{acc}} 	& \diagbox{*\tsc{nom}}{*\tsc{gen}} 	\\\hline
		V\scsub{acc} 		& \diagbox{\tsc{acc}}{*\tsc{nom}} & \tsc{acc} 												& \diagbox{*\tsc{acc}}{*\tsc{gen}} 	\\\hline
		V\scsub{gen} 		& \diagbox{\tsc{gen}}{*\tsc{nom}} & \diagbox{*\tsc{gen}}{*\tsc{acc}}	& \tsc{gen}													\\\hline
	\end{tabular}
\end{table}

\ex. after inserting clitics
\ag. Me efχarístisan ópji tus íχa δósi leftá.\\
\tsc{cl.1sg.acc} thanked.3\tsc{pl}\scsub{nom} who.\tsc{nom.pl} \tsc{cl.3pl.gen} had.1\tsc{sg}\scsub{gen} given money\\
`Whoever I had given money to thanked me.' \hfill m:\tsc{nom}, e:\tsc{gen}
\bg. Γnórisa ópjon tu éδosan tin ipotrofia.\\
met.1\tsc{sg}\scsub{acc} who.\tsc{acc.sg} \tsc{cl.3sg.gen} gave.3\tsc{pl}\scsub{gen} the scholarship.\tsc{acc}\\
`I met whoever they gave the scholarship to.' \hfill m:\tsc{acc}, e:\tsc{gen}
\b. Lefteris?
\gll Eδósa leftá ópju ton γnórisa.\\
gave.1\tsc{sg}\scsub{gen} money who.\tsc{gen.sg} \tsc{cl.3sg.acc} met.1\tsc{sg}\scsub{acc}\\
`I gave money to whoever I met.' \hfill m:\tsc{gen}, e:\tsc{acc}



\begin{table}[h]
	\center
	\setlength{\tabcolsep}{10pt}
	\begin{tabular}{|c|c|c|c|}\hline
		\diagbox{m}{e}	& V\scsub{nom} 										& V\scsub{acc} 											& V\scsub{gen} 																					\\\hline
		V\scsub{nom} 		& \tsc{nom} 											& \diagbox{\tsc{nom}}{*\tsc{acc}}		& \diagbox{\tsc{nom} \tiny{+\tsc{cl.gen}}}{*\tsc{gen}} 	\\\hline
		V\scsub{acc} 		& \diagbox{\tsc{acc}}{*\tsc{nom}} & \tsc{acc} 												& \diagbox{\tsc{acc} \tiny{+\tsc{cl.gen}}}{*\tsc{gen}} 	\\\hline
		V\scsub{gen}	  & \diagbox{\tsc{gen}}{*\tsc{nom}} & \diagbox{*\tsc{gen}}{*\tsc{acc}}	& \tsc{gen}																							\\\hline
	\end{tabular}
\end{table}
